Discovering Root Causal Genes with High Throughput Perturbations

Version 1

Abstract

Abstract

Root causal gene expression levels – or root causal genes for short – correspond to the initial changes to gene expression that generate patient symptoms as a downstream effect. Identifying root causal genes is critical towards developing treatments that modify disease near its onset, but no existing algorithms attempt to identify root causal genes from data. RNA-sequencing (RNA-seq) data introduces challenges such as measurement error, high dimensionality and non-linearity that compromise accurate estimation of root causal effects even with state-of-the-art approaches. We therefore instead leverage Perturb-seq, or high throughput perturbations with single cell RNA-seq readout, to learn the causal order between the genes. We then transfer the causal order to bulk RNA-seq and identify root causal genes specific to a given patient for the first time using a novel statistic. Experiments demonstrate large improvements in performance. Applications to macular degeneration and multiple sclerosis also reveal root causal genes that lie on known pathogenic pathways, delineate patient subgroups and implicate a newly defined omnigenic root causal model.

Introduction

Genetic and non-genetic factors that cause disease can affect gene expression as an intermediate step. We introduce root causal gene expression levels – or root causal genes for short – that correspond to the initial changes to gene expression induced by genetic and non-genetic factors that generate a pathogenic cascade ultimately causing a downstream diagnosis (Figure 1). Root causal genes differ from core genes that directly affect the diagnosis and thus lie at the end, rather than at the beginning, of pathogenesis (Boyle et al., 2017). Root causal genes also generalize (the expression levels of) driver genes that only account for the effects of somatic mutations primarily in cancer (Martínez-Jiménez et al., 2020).

Toy example where a variable E2 simultaneously models genetic and non-genetic factors that cause a diagnose Y through gene expression . In this case, E2 first affects the gene expression level , or the root causal gene. The root causal gene then affects other downstream levels during pathogenesis, ultimately inducing a diagnosis Y.

Treating root causal genes can modify disease pathogenesis in its entirety, whereas targeting other causes may only provide symptomatic relief. For example, mutations in Gaucher disease cause decreased expression of wild type beta-glucocerebrosidase, or the root causal gene (Nagral, 2014). We can give a patient blood transfusions to alleviate the fatigue and anemia associated with the disease, but we seek more definitive treatments like recombinant glucocerebrosidase that replaces the deficient enzyme. Enzyme replacement therapy alleviates the associated liver, bone and neurological abnormalities of Gaucher disease as a downstream effect. Identifying root causal genes is therefore critical for developing treatments that eliminate disease near its pathogenic onset.

The problem is further complicated by the existence of complex disease, where a patient may have multiple root causal genes that differ from other patients even within the same diagnostic category (Cano-Gamez and Trynka, 2020). Complex diseases often have an overwhelming number of causes, but the root causal genes may only represent a small subset implicating a more omnigenic than polygenic model. We thus also seek to identify patient-specific root causal genes in order to classify patients into meaningful biological subgroups each hopefully dictated by only a small group of genes.

No existing method identifies root causal genes from data. Many algorithms focus on discovering associational (Costa-Silva et al., 2017) or even causal relations (Wang et al., 2023; Wen et al., 2023), but none pinpoint the first gene expression levels that ultimately generate disease. We therefore define the Root Causal Strength (RCS) score to identify root causal genes unique to each patient. We then design the Root Causal Strength using Perturbations (RCSP) algorithm that estimates RCS from bulk RNA-seq under minimal assumptions by integrating Perturb-seq, or high throughput perturbation experiments using CRISPR-based technologies coupled with single cell RNA-sequencing (Dixit et al., 2016; Adamson et al., 2016; Datlinger et al., 2017). Experiments demonstrate marked improvements in performance. Finally, application of the algorithm to two complex diseases with disparate pathogeneses simultaneously recovers root causal genes, omnigenic disease models, pathogenic pathways and drug candidates delineating patient subgroups for the first time.

Results

We briefly summarize the Methods in the first two subsections.

Definitions

Differential expression analysis identifies differences in gene expression levels between groups Y (Costa-Silva et al., 2017). A gene Xi may be differentially expressed due to multiple reasons. For example, Xi may cause Y, or a confounder C may explain the relation between Xi and Y such that Xi ← C → Y. In this paper, we take expression analysis a step further by pinpointing causal relations from expression levels regardless of the variable type of Y (discrete or continuous). We in particular seek to discover patient-specific root causal genes from bulk RNA-seq data, which we carefully define below.

We represent a biological system in bulk RNA-seq as a causal graph 𝔾 – such as in Figure 2 (a) – where p vertices represent true gene expression levels in a bulk sample and Y denotes the patient symptoms or diagnosis. The set contains thousands of genes in practice. Directed edges between the vertices in 𝔾 refer to direct causal relations. We assume that gene expression causes patient symptoms but not vice versa so that no edge from Y is directed towards . The set refers to the parents of , or those variables with an edge directed into . For example, in Figure 2 (a). A root vertex corresponds to a vertex with no parents.

Method overview and synthetic data results. (a) We consider a latent causal graph over the true counts . (b) We augment the graph with error terms E such that each Ei ∈ E in red has an edge directed towards . (c) The RCS of , denoted by Φ2, quantifies the strength of the root causal effect from E to Y conditional on . (d) We cannot observe in practice but instead observe the noisy surrogates X in blue corrupted by Poisson measurement error. (e) Perturbing a variable such as changes the marginal distributions of downstream variables shown in green under mild conditions. (f) RCSP thus uses the perturbation data to identify the set of surrogate parents for each variable in order to compute Φ. (g) Violin plots show that RCSP achieved the smallest RMSE to the ground truth RCS values in the synthetic data. (h) RCSP also took about the same amount of time to complete as multivariate regression. Univariate regression only took 11 seconds on average, so its bar is not visible. Error bars denote 95% confidence intervals of the mean. (i) Finally, RCSP maintained low RMSE values regardless of the number of clusters considered.

We can associate 𝔾 with the structural equation for each that links each vertex to its parents and error term Ei (Pearl, 2009). The error term Ei is not simply a regression residual but instead represents the conglomeration of unobserved explanatory variables that only influence , such as unobserved transcriptional regulators, certain genetic variants and specific environmental conditions. We thus also include the error terms E in the directed graph of Figure 2 (b). All root vertices are error terms and vice versa. The root causes of Y are the error terms that cause Y, or have a directed path into Y. We define the root causal strength (RCS) of on Y as the following absolute difference (Figure 2 (c)):

We prove the last equality in the Methods. As a result, RCS Φi directly measures the contribution of the gene on Y according to its error term Ei without recovering the error term values. The algorithm does not impose distributional assumptions or functional restrictions such as additive noise to estimate the error term values as an intermediate step. Moreover Φi is patient-specific because the values of and may differ between patients. We have Φi = 0 when Ei is not a cause of Y, so we say that the gene is a patient-specific root causal gene if Φi > 0.

Algorithm

We propose an algorithm called Root Causal Strength using Perturbations (RCSP) that estimates Φ = {Φ1, …, Φp} from genes measured in both bulk RNA-seq and Perturb-seq datasets derived from possibly independent studies but from the same tissue type. We rely on bulk RNA-seq instead of single cell RNA-seq in order to obtain many samples of the label Y. We focus on statistical estimation rather than statistical inference because Φi > 0 when when Ei causes Y under mild conditions, so we reject the null hypothesis that Φi = 0 for many genes if many gene expression levels cause Y. However, we hypothesize that Φi ≫ 0 for a small subset of genes in each patient under an omnigenic model, which we seek to estimate.

Estimating Φ requires access to the true gene expression levels and the removal of the effects of confounding. We first control for batch effects representing unwanted sources of technical variation such as different sequencing platforms or protocols. We however can only obtain imperfect counts X from RNA sequencing even within each batch (Figure 2 (d)). Measurement error introduces confounding as well because it prevents us from exactly controlling for the causal effects of the gene expression levels. Investigators usually mitigate measurement error by normalizing the gene expression levels by sequencing depth. We show in the Methods that the Poisson distribution approximates the measurement error distribution induced by the sequencing process to high accuracy (Choudhary and Satija, 2022; Sarkar and Stephens, 2021). We leverage this fact to eliminate the need for normalization by sequencing depth using an asymptotic argument where the library size N approaches infinity. N takes on a value of at least ten million in bulk RNA-seq, but we also empirically verify that the theoretical results hold well in the Supplementary Materials. We thus eliminate the Poisson measurement error and batch effects by controlling for the batches B but not N in non-linear regression models.

We in particular show that Φi in Equation (1) is also equivalent to: where refers to the surrogate parents of , or the variables in X associated with . RCSP can identify (an appropriate superset of) the surrogate parents of each variable using perturbation data because perturbing a gene changes the marginal distributions of its downstream effects – which the algorithm detects from data under mild assumptions (Figures 2 (e) and (f)). The algorithm thus only transfers the binary presence or absence of causal relations from the single cell to bulk data – rather than the exact functional relationships – in order to remain robust against discrepancies between the two data types. RCSP finally performs the two non-linear regressions needed to estimate and for each Φi. We will compare Φi against Statistical Dependence (SD), a measure of correlational strength defined as Ωi = | 𝔼(Y| Xi, B) − 𝔼(Y | B) | where we have removed the conditioning on .

In silico identification of root causal genes

We simulated 30 bulk RNA-seq and Perturb-seq datasets from random directed graphs summarizing causal relations between gene expression levels. We performed single gene knock-down perturbations over 2500 genes and 100 batches. We obtained 200 cell samples from each perturbation, and another 200 controls without perturbations. We therefore generated a total of 2501 × 200 = 500, 200 single cell samples for each Perturb-seq dataset. We simulated 200 bulk RNA-seq samples. We compared RCSP against the Additive Noise Model (ANM) (Peters et al., 2014; Strobl and Lasko, 2023a), the Linear Non-Gaussian Acyclic Model (LiNGAM) (Peters et al., 2014; Strobl and Lasko, 2022), CausalCell (Wen et al., 2023), univariate regression residuals (Uni Reg), and multivariate regression residuals (Multi Reg). The first two algorithms are state-of-the-art approaches used for error term extraction and, in theory, root causal discovery. See Methods for comprehensive descriptions of the simulation setup and comparator algorithms.

We summarize accuracy results in Figure 2 (g) using the Root Mean Squared Error (RMSE) to the ground truth Φ values. All statements about pairwise differences hold true at a Bonferonni corrected threshold of 0.05/5 according to paired two-sided t-tests, since we compared a total of five algorithms. RCSP estimated Φ most accurately by a large margin. ANM and LiNGAM are theoretically correct under their respective assumptions, but they struggle to outperform standard multivariate regression due to the presence of measurement error in RNA-seq (Supplementary Materials). Feature selection and causal discovery with CausalCell did not improve performance. Univariate regression performed the worst, since it does not consider the interactions between variables. RCSP achieved the lowest RMSE while completing in about the same amount of time as multivariate regression on average (Figure 2 (h)). We conclude that RCSP both scalably and accurately estimates Φ from a combination of bulk RNA-seq and Perturb-seq data.

We will cluster the RCS values in real data to find patient subgroups. We therefore also performed hierarchical clustering using Ward’s method (Ward Jr, 1963) on the values of Φ estimated by RCSP with the synthetic data. We then computed the RMSE within each cluster and averaged the RMSEs between clusters. We found that RCSP maintained low average RMSE values regardless of the number of clusters considered (Figure 2 (i)). We conclude that RCSP maintains accurate estimation of Φ across different numbers of clusters.

Oxidative stress in age-related macular degeneration

We ran RCSP on a bulk RNA-seq dataset of 513 individuals with age-related macular degeneration (AMD; GSE115828) and a Perturb-seq dataset of 247,914 cells generated from an immortalized retinal pigment epithelial (RPE) cell line (Ratnapriya et al., 2019; Replogle et al., 2022). The Perturb-seq dataset contains knockdown experiments of 2077 genes overlapping with the genes of the bulk dataset. We set the target Y to the Minnesota Grading System score, a measure of the severity of AMD based on stereoscopic color fundus photographs. We always included age and sex as a biological variable as covariates. We do not have access to the ground truth values of Φ in real data, so we evaluated RCSP using seven alternative techniques. See Methods for a detailed rationale of the evaluation of real data. RCSP outperformed all other algorithms in this dataset (Supplementary Materials). We therefore only analyze the output of RCSP in detail here.

AMD is a neurodegenerative disease of the aging retina (Hadziahmetovic and Malek, 2021), so age is a known root cause of the disease. We therefore determined if RCSP identified age as a root cause. Note that RCSP does not need perturbation data of age to compute the RCS values of age, since age has no parents in the directed graph. The algorithm estimated a heavy tailed distribution of the RCS values indicating that most of the RCS values deviated away from zero (Figure 3 (a)). The Deviation of the RCS (D-RCS), or the standard deviation from an RCS value of zero, measures the tailedness of the distribution while preserving the unit of measurement. The D-RCS of age corresponded to 0.46 – more than double that of the nearest gene (Figure 3 (d)). We conclude that RCSP correctly detected age as a root cause of AMD.

Analysis of AMD. (a) The distribution of the RCS scores of age deviated away from zero and had a composite D-RCS of 0.46. (b) However, the majority of gene D-RCS scores concentrated around zero, whereas the majority of gene D-SD scores concentrated around the relatively larger value of 0.10. Furthermore, the D-RCS scores of the genes in (d) mapped onto the “amino acid transport across the plasma membrane” pathway known to be involved in the pathogenesis of AMD in (c). Blue bars survived 5% FDR correction. (e) Drug enrichment analysis revealed four significant drugs, the later three of which have therapeutic potential. (f) Hierarchical clustering revealed four clear clusters according to the elbow method, which we plot by UMAP dimensionality reduction in (g). The RCS scores of the top genes in (d) increased only from the left to right on the first UMAP dimension (x-axis); we provide an example of SLC7A5 in (h) and one of three detected exceptions in (i). We therefore performed pathway enrichment analysis on the black cluster in (g) containing the largest RCS scores. (j) The amino acid transport pathway had a larger degree of enrichment in the black cluster as compared to the global analysis in (c).

Root causal genes typically affect many downstream genes before affecting Y. We therefore expect to identify few root causal genes but many genes that correlate with Y. To evaluate this hypothesis, we examined the distribution of D-RCS relative to the distribution of the Deviation of Statistical Dependence (D-SD), or the standard deviation from an SD value of zero, in Figure 3 (b). Few D-RCS scores had large values implying the existence of only a few significant root causal genes. In contrast, most of the D-SD scores had relatively larger values concentrated around 0.10 implying the existence of many genes correlated with Y. We conclude that RCSP identified few root causal genes rather than many correlated genes for AMD.

The pathogenesis of AMD involves the loss of RPE cells. The RPE absorbs light in the back of the retina, but the combination of light and oxygen induces oxidative stress, and then a cascade of events such as immune cell activation, cellular senescence, drusen accumulation, neovascularization and ultimately fibrosis (Barouch and Miller, 2007). We therefore expect the root causal genes of AMD to include genes involved in oxidative stress. The gene MIPEP with the highest D-RCS score in Figure 3 (d) indeed promotes the maturation of oxidative phosphorylation-related proteins (Shi et al., 2011). The second gene SLC7A5 is a solute carrier that activates mTORC1 whose hyperactivation increases oxidative stress via lipid peroxidation (Nachef et al., 2021; Go et al., 2020). The gene HEATR1 is involved in ribosome biogenesis that is downregulated by oxidative stress (Turi et al., 2018). The top genes discovered by RCSP thus identify pathways known to be involved in oxidative stress.

We subsequently jointly analyzed the D-RCS values of all 2077 genes. We performed pathway enrichment analysis that yielded one pathway “amino acid transport across the plasma membrane” that passed an FDR threshold of 5% (Figure 3 (c)). The leading edge genes of the pathway included the solute carriers SLC7A5 and SLC1A5. These two genes function in conjunction to increase the efflux of essential amino acids out of the lysosome (Nicklin et al., 2009; Beaumatin et al., 2019). Some of these essential amino acids like L-leucine and L-arginine activate mTORC1 that in turn increases lipid peroxidation induced oxidative stress and the subsequent degeneration of the RPE (Nachef et al., 2021; Go et al., 2020). We conclude that pathway enrichment analysis correctly identified solute carrier genes involved in a known pathway promoting oxidative stress in AMD.

We next ran drug enrichment analysis with the D-RCS scores. The top compound arsenous acid inhibits RPE proliferation (Su et al., 2020), but the other three significant drugs have therapeutic potential (Figure 3 (e)). Busulfan decreases the requirement for intravitreal anti-VEGF injections (Dalvin et al., 2022). Genistein is a protein kinase inhibitor that similarly attenuates neovascularization (Kinoshita et al., 2014) and blunts the effect of ischemia on the retina (Kamalden et al., 2011). Finally, a metabolite of the antiviral agent 3’-azido-3’-deoxythymidine inhibits neovascularization and mitigates RPE degeneration (Narendran et al., 2020). We conclude that the D-RCS scores identified promising drugs for the treatment of AMD.

Hierarchical clustering and UMAP dimensionality reduction on the patient-specific RCS values revealed four clear clusters of patients by the elbow method on the sum of squares plot (Figures 3 (f) and (g), respectively). Most of the top genes exhibited a clear gradation increasing only from the left to the right hand side of the UMAP embedding; we plot an example in Figure 3 (h). We found three exceptions to this rule among the top 30 genes (example in Figure 3 (i) and see Supplementary Materials). RCSP thus detected genes with large RCS scores primarily in the black cluster of Figure 3 (g). Pathway enrichment analysis within this cluster alone yielded supra-significant results on the same pathway detected in the global analysis (Figure 3 (j) versus Figure 3 (c)). Furthermore, clusterwise drug enrichment analysis results confirmed that patients in the black cluster with many root causal genes are likely the hardest to treat (Supplementary Materials). We conclude that RCSP detected a subgroup of patients whose root causal genes have large RCS scores and involve known pathogenic pathways related to oxidative stress.

T cell infiltration in multiple sclerosis

We next ran RCSP on 137 samples collected from CD4+ T cells of multiple sclerosis (MS; GSE137143) as well as Perturb-seq data of 1,989,578 lymphoblasts, or the precursors of T cells and other lymphocytes (Kim et al., 2021; Replogle et al., 2022). We set the target Y to the Expanded Disability Status Scale score, a measure of MS severity. RCSP outperformed all other algorithms in this dataset as well (Supplementary Materials).

MS progresses over time, and RCSP correctly detected age as a root cause of MS severity with RCS values deviating away from zero (Figure 4 (a)). The distribution of gene D-RCS scores concentrated around zero, whereas the distribution of gene D-SD scores concentrated around a relatively larger value of 0.3 (Figure 4 (b)). RCSP thus detected an omnigenic root causal model rather than a polygenic correlational model.

Analysis of MS. (a) The distribution of the RCS scores of age deviated away from zero with a composite D-RCS of 0.55. (b) The distribution of D-RCS concentrated around zero, whereas the distribution of D-SD concentrated around 0.3. (d) RCSP identified many genes with large D-RCS scores that in turn mapped onto known pathogenic pathways in MS in (c). Hierarchical clustering revealed three clusters in (e), which we plot in two dimensions with UMAP in (f). Top genes did not correlate with either dimension of the UMAP embedding; we provide an example of the MNT gene in (g). (h) Drug enrichment analysis in the green cluster implicated multiple cathepsin inhibitors. Finally, EPH-ephrin signaling survived FDR correction in (c) and was enriched in the pink cluster in (i) which contained more MS patients with the relapsing-remitting subtype in (j); subtypes include relapse-remitting (RR), primary progressive (PP), secondary progressive (SP), clinically isolated syndrome (CIS), and radiologically isolated syndrome (RIS).

MS is an inflammatory neurodegenerative disease that damages the myelin sheaths of nerve cells in the brain and spinal cord. T cells may mediate the inflammatory process by crossing a disrupted blood brain barrier and repeatedly attacking the myelin sheaths (Fletcher et al., 2010). Damage induced by the T cells also perturbs cellular homeostasis and leads to the accumulation of misfolded proteins (Andhavarapu et al., 2019). The root causal genes of MS thus likely include genes involved in T cell infiltration across the blood brain barrier.

Genes with the highest D-RCS scores included MNT, CERCAM and HERPUD2 (Figure 4 (d)). MNT is a MYC antagonist that modulates the proliferative and pro-survival signals of T cells after engagement of the T cell receptor (Gnanaprakasam and Wang, 2017). Similarly, CERCAM is an adhesion molecule expressed at high levels in microvessels of the brain that increases leukocyte transmigration across the blood brain barrier (Starzyk et al., 2000). HERPUD2 is involved in the endoplasmic-reticulum associated degradation of unfolded proteins (Kokame et al., 2000). Genes with the highest D-RCS scores thus serve key roles in known pathogenic pathways of MS.

We found multiple genes with high D-RCS scores in MS, in contrast to AMD where age dominated (Figure 4 (d) versus Figure 3 (d)). We performed pathway enrichment analysis using the D-RCS scores of all genes and discovered two significant pathways at an FDR corrected threshold of 5%: “adenomatous polyposis coli (APC) truncation mutants have impaired AXIN binding” and “EPH-ephrin signaling” (Figure 4 (c)). APC and AXIN are both members of the Wnt signaling pathway and regulate levels of beta-catenin (Spink et al., 2000). Furthermore, inhibition of Wnt/beta-catenin causes CD4+ T cell infiltration into the central nervous system via the blood brain barrier in MS (Lengfeld et al., 2017). Ephrins similarly regulate T cell migration into the central nervous system (Luo et al., 2016) and are overexpressed in MS lesions (Sobel, 2005). The APC-AXIN and EPH-ephrin pathways are thus consistent with the known pathophysiology of central nervous system T cell infiltration in MS.

We subsequently performed hierarchical clustering of the RCS scores. The within cluster sum of squares plot in Figure 4 (e) revealed the presence of three clusters by the elbow method. We plot the three clusters in a UMAP embedding in Figure 4 (f). The clusters did not show a clear relationship with MS symptom severity (Supplementary Materials) or the levels of the top most genes of Figure 4 (d); we plot the MNT gene as an example in Figure 4 (g). However, further analyses with additional genes revealed that the distribution of many lower ranked genes governed the structure of the UMAP embedding (Supplementary Materials). The D-RCS scores of each cluster also implicated different mechanisms of T cell pathology including APC-AXIN in the green cluster, disturbed T cell homeostasis in the pink cluster and platelet enhanced T cell autoreactivity in the blue cluster (Supplementary Materials).

Global drug enrichment analysis did not yield any significant drugs even at a liberal FDR threshold of 10%. We thus ran drug enrichment analysis in each cluster of Figure 4 (f). The blue and pink clusters again did not yield significant drugs. However, the third green cluster identified the cysteine cathepsin inhibitors dipeptide-derived nitriles, phenylalinine derivatives, e-64, L-006235 and L-873724 (Figure 4 (h)); statistical significance of the first three held even after correcting for multiple comparisons with the Bonferroni adjustment of 0.05/4 on the q-values. The leading edge genes of the significant drugs included the cathepsins CTSL, CTSS and CTSB exclusively. These drug enrichment results corroborate multiple experimental findings highlighting the therapeutic efficacy of cathepsin inhibitors in a subgroup of MS patients responsive to interferon therapy (Haves-Zburof et al., 2011; Burster et al., 2007).

Prior research has also shown that EPH-ephrin signaling is more prevalent in relapsing-remitting multiple sclerosis than in other subtypes of the disease (Golan et al., 2021). EPH-ephrin signaling survived FDR correction in our analysis (Figure 4 (c)). Furthermore, the pathway was more enriched in the pink cluster than in the other two (Figure 4 (i)). The pink cluster indeed contained a higher proportion of patients with the relapsing-remitting subtype (Figure 4 (j)). RCSP thus precisely identified the enrichment of EPH-ephrin signaling in the correct subtype of MS.

Discussion

We presented a framework for identifying root causal genes using the error terms of structural equation models. Each error term represents the conglomeration of unobserved root causes, such as genetic variants or environmental conditions, that only modulate a specific gene. We however do not have access to many of the error terms in practice, so we introduced the root causal strength (RCS) score that detects root causal genes from bulk RNA-seq without recovering the error term values as an intermediate step. The RCSP algorithm computes RCS given knowledge of the causal ancestors of each variable, which we obtained by Perturb-seq. RCSP only transfers the causal structure (binary cause-effect relations) from the single cell to bulk data rather than the exact functional relationships in order to remain robust against discrepancies between the two data types. Results with the synthetic data demonstrated marked improvements over existing alternatives. The algorithm also recovered root causal genes that play key roles in known pathogenic pathways and implicate therapeutic drugs in both AMD and MS.

We detected a modest number of root causal genes in both AMD and MS relative to the number of genes correlated with Y. This omnigenic model differs from the omnigenic model involving core genes (Boyle et al., 2017). Boyle et al. define core genes as genes that directly affect disease risk and play specific roles in disease etiology. In contrast, root causal genes may not directly affect Y but lie substantially upstream of Y in the causal graph. Boyle et al. further elaborate that many peripheral genes affect the functions of a modest number of core genes, so the peripheral genes often explain most of disease heritability. Causation from root causal genes moves in the opposite direction – the error terms of upstream root causal genes causally affect many downstream genes that include both ancestors and non-ancestors of Y. These downstream genes contain traces of the root causal gene error terms that induce the many correlations with Y. The error terms of root causal genes associated with large RCS scores also mix with the error terms of the other ancestors of Y with small RCS scores leading to Fisher’s classic infinitesimal model (Fisher, 1919). The indirect effects of root causal genes on Y and the impact of root causal genes on many downstream genes correlated with Y motivate us to use the phrase omnigenic root causal model in order to distinguish it from the omnigenic core gene model.

We identified root causal genes without imposing parametric assumptions using the RCS metric. Prior measures of root causal effect require restrictive functional relations, such as linear relations or additive noise, and continuous random variables (Strobl and Lasko, 2022; Strobl et al., 2023; Strobl and Lasko, 2023a). These restrictions ensure exact identifiability of the underlying causal graph and error terms. However, real RNA-seq is obtained from a noisy sequencing process and contains count data arguably corrupted by Poisson measurement error (Sarkar and Stephens, 2021). The Poisson measurement error introduces confounding that precludes exact recovery of the underlying error terms. The one existing root causal discovery method that can handle Poisson measurement error uses single cell RNA-seq, estimates negative binomial distribution parameters and cannot scale to the thousands of genes required for meaningful root causal detection (Strobl and Lasko, 2023b). RCSP rectifies the deficiencies of these past approaches by ensuring accurate root causal detection even in the presence of the counts, measurement error and high dimensionality of RNA-seq.

This study carries limitations worthy of addressing in future work. The RCS score importantly quantifies root causal strength rather than root causal effect. As a result, the method cannot be used to identify the direction of root causal effect unconditional on the parents. The root causal effect and RCS do not differ by much in practice (Supplementary Materials), but future work should focus on exactly identifying both the strength and direction of the causal effects of the error terms. Furthermore, RCS achieves patient but not cell-specificity because the algorithm relies on phenotypic labels obtained from bulk RNA-seq. RCSP thus cannot identify the potentially different root causal genes present within distinct cell populations. Third, RCSP accounts for known batch effects and measurement error but cannot adjust for unknown confounding. Finally, RCSP assumes a directed acyclic graph. We can transform a directed graph with cycles into an acyclic one under equilibrium, but real biological distributions vary across time (Spirtes, 1995; Bongers et al., 2021). Future work should thus aim to estimate cell-specific root causal effects under latent confounding and time-varying distributions.

In conclusion, RCSP integrates bulk RNA-seq and Perturb-seq to identify patient-specific root causal genes under a principled causal inference framework using the RCS score. RCS quantifies root causal strength implicitly without requiring normalization by sequencing depth or direct access to the error terms of a structural equation model. The algorithm identifies the necessary causal relations to compute RCS using reliable high throughput perturbation data rather than observational data alone. The RCS scores often suggest an omnigenic rather than a polygenic root causal model of disease. Enrichment analyses with the RCS scores frequently reveal pathogenic pathways and drug candidates. We conclude that RCSP is a novel, accurate, scalable and disease-agnostic procedure for performing patient-specific root causal discovery.

Methods and Materials

Background on Causal Discovery

We denote a singleton variable like with italics and sets of variables like with bold italics. We can represent a causal process using a structural equation model (SEM) linking the p + 1 variables in using a series of deterministic functions: where fi is a function of the parents, or direct causes, of Zi and an error term Ei ∈ E. The error terms E are mutually independent. We will use the terms vertex and variable interchangeably. A root vertex corresponds to a vertex without any parents. On the other hand, a terminal vertex is not a parent of any other vertex.

We can associate a directed graph to Z by drawing a directed edge from each member of Pa(Zi) to Zi for all Zi ∈ Z. A directed path from Zi to Zj corresponds to a sequence of adjacent directed edges from Zi to Zj. If such a path exists (or Zi = Zj), then Zi is an ancestor of Zj and Zj is a descendant of Zi. We collate all ancestors of Zi into the set Anc(Zi). A cycle occurs when there exists a directed path from Zi to Zj and the directed edge Zj → Zi. A directed acyclic graph (DAG) contains no cycles. We augment a directed graph by including additional vertices E and drawing a directed edge from each Ei ∈ E to Xi except when Xi = Ei is already a root vertex. We consider an augmented DAG 𝔾 throughout the remainder of this manuscript.

The vertices Zi and Zj are d-connected given W ⊆ Z \{Zi, Zj } in 𝔾 if there exists a path between Zi and Zj such that every collider on the path is an ancestor of W and no non-collider is in W. The vertices are d-separated if they are not d-connected. Any DAG associated with the SEM in Equation (3) also obeys the global Markov property where Zi and Zj are conditionally independent given W if they are d-separated given W. The term d-separation faithfulness refers to the converse of the global Markov property where conditional independence implies d-separation. A distribution obeys unconditional d-separation faithfulness when we can only guarantee d-separation faithfulness when W = ∅.

Causal Modeling of RNA Sequencing

Performing causal discovery requires careful consideration of the underlying generative process. We therefore propose a causal model for RNA-seq. We differentiate between the biology and the RNA sequencing technology.

We represent a biological causal process using an SEM over obeying Equation (3). We assume that the phenotypic target Y is a terminal vertex so that gene expression causes phenotype but not vice versa. Each corresponds to the total number of RNA molecules of a unique gene in a single cell or bulk tissue sample. We unfortunately cannot observe in practice but instead measure a corrupted count X using single cell or bulk RNA-seq technology.

We derive the measurement error distribution from first principles. We map an exceedingly small fraction of each within a sample at unequal coverage. Let πij denote the probability of mapping one molecule of in batch j so that is near zero. The law of rare events (Papoulis, 1984) implies that the Poisson distribution well-approximates the library size N so that .

We write the probability of mapping in a given sample as:

This proportion remains virtually unchanged when sampling without replacement because with small . We can therefore approximate sampling without replacement by sampling with replacement using a multinomial:. This multinomial and the Poisson distribution over N together imply that the marginal distribution of each Xi ∈ X follows an independent Poisson distribution centered at , or:

We conclude that the measurement error distribution follows a Poisson distribution to high accuracy. Multiple experimental results already corroborate this theoretical conclusion (Grün et al., 2014; Sarkar and Stephens, 2021; Choudhary and Satija, 2022).

We can represent the biology and the RNA sequencing in a single DAG over , where B denotes the batch, and Y the target variable representing patient symptoms or diagnosis. We provide a toy example in Figure 5. We draw 𝔾 over Z in black and make each a parent of Xi ∈ X in blue. We then include the root vertex B as a parent of all members of X in green. We augment this graph with the error terms of in red and henceforth refer to the augmented DAG as 𝔾. Repeated draws from the represented causal process generates a dataset.

An example of a DAG over augmented with the error terms E. The observed vertices X denote counts corrupted by batch B effects and Poisson measurement error.

No Need for Normalization by Sequencing Depth

We provide an asymptotic argument that eliminates the need for normalization by sequencing depth when estimating conditional expectations using bulk RNA-seq. The argument applies to the conditional expectations as a whole rather than their individual parameters.

We want to recover the causal relations between by removing batch B and depth N effects from the dataset because they correspond to the sequencing process rather than the underlying biology. We first consider removing sequencing depth by finding stably expressed housekeeping genes. Let denote the set of housekeeping genes where is a constant for each ; similarly A refers to the corresponding set with Poisson measurement error. Let N = n be large enough such that for each sample. Then dividing by controls for sequencing depth in the following sense: where we have divided by a constant in the last term. Thus, dividing by L removes measurement error within each batch as N → ∞. We assume that N is so large that the approximation error is negligible. We only invoke the assumption in bulk RNA-seq, where the library size N is on the order of at least tens of millions.

We do not divide by L in practice because we may have L = 0 with finite N. We instead always include L ∪ B in the predictor set of downstream regressions. Conditioning on L ∪ B ensures that all downstream regressions mitigate depth and batch effects with adequate sequencing depth, or that for any as N → ∞. The equality holds almost surely under a mild smoothness condition:

Assume Lipschitz continuity of the conditional expectation for all N ≥ n0 : where is a positive constant, and we have taken an outer expectation on both sides. Then almost surely.

We delegate proofs to the Supplementary Materials unless proven here in the Methods. Note that , so the Lipschitz assumption intuitively means that accurate estimation of implies accurate estimation of . Furthermore, conditioning on the library size N instead of L can introduce spurious dependencies because N depends on all of the genes rather than just the stably expressed ones.

We now eliminate the need to condition on L. Note that L is a sum of independent Poisson distributions given B per Expression (4). This implies for any N, so that almost surely. We have proved:

Consider the same assumption as Lemma 1. Then almost surely, where we have eliminated the conditioning on L.

We emphasize again that these equalities hold for the conditional expectation but not for the regression parameters; the regression parameters do not converge in general unless we divide by L. We will only need to estimate conditional expectations in order to identify root causal genes.

Identifying Root Causal Genes

We showed how to overcome Poisson measurement error without sequencing depth normalization in the previous section. We leverage this technique to define a measure for identifying the root causal genes of Y.

Definitions

A root cause of Y corresponds to a root vertex that is an ancestor of Y in 𝔾. All root vertices are error terms in an augmented graph. We define the root causal effect of any Ei∈ E on Y as γi ≜ ℙ(Y| Ei) − ℙ(Y) (Strobl, 2024; Strobl and Lasko, 2023c).

We can identify root causes using the following result:

If or (or both), then Ei is a root cause of Y.

We can also claim the backward direction under d-separation faithfulness. We however avoid making this additional assumption because real biological data may not arise from distributions obeying d-separation faithfulness in practice (Strobl, 2022).

Proposition 1 implies that Ei is a root cause of Y when:

However, Δi does not correspond to the root causal effect γi due to the extra conditioning on . The two terms may also differ in direction; if Δ > 0, then this does not imply that γ > 0, and similarly for negative values. The two variables thus represent different quantities but – in terms of priority – we would estimate γi when we have nonzero Δi. Experimental results indicate that γi and Δi take on similar values and agree in direction about 95% of the time in practice (Supplementary Materials).

We now encounter two challenges. First, the quantities γi and Δi depend on the unknown error term Ei. We can however substitute Ei with in Δi due to the following result:

We have under Equation (3).

We can thus compute Δi without access to the error terms:

The ability to determine the root causal status of Ei on Y when Δi ≠ 0 per Proposition 1, and the above ability to directly substitute Ei with its gene both motivate the following definition: we say that is a root causal gene of Y if Δi ≠ 0.

The second challenge involves computing the non-parametric probability distributions of Δi which come at a high cost. We thus define the analogous expected version by: where p(Y) denotes the density of Y. Observe that if Δi = 0, then Γi = 0. The converse is not true but likely to hold in real data when a change in the probability distribution also changes its expectation. The set denotes the surrogate parents of corresponding to the variables in X associated with . The last equality holds almost surely as N → ∞ by Theorem 1.

We call Φi ≜ |Γi| the Root Causal Strength (RCS) of on Y. The RCS obtains a unique value Φi = ϕij for each patient j. We say that is a root causal gene of Y for patient j if its RCS score is non-zero. We combine the RCS scores across a set of n samples using the Deviation of the RCS , or the standard deviation of RCS from zero. We may compute D-RCS for each cluster or globally across all patients depending on the context. We thus likewise say that is a root causal gene for a cluster of patients or all patients in a sample if its corresponding D-RCS score for the cluster or the sample is non-zero, respectively.

Algorithm

We now design an algorithm called Root Causal Strength using Perturbations (RCSP) that recovers the RCS scores using Perturb-seq and bulk RNA-seq data.

Finding Surrogate Ancestors

Computing Φi for each requires access to the surrogate parents of each variable or, equivalently, the causal graph 𝔾. However, inferring 𝔾 using causal discovery algorithms may lead to large statistical errors in the high dimensional setting (Colombo et al., 2014) and require restrictive assumptions such as d-separation faithfulness (Spirtes et al., 2000) or specific functional relations (Peters et al., 2014).

We instead directly utilize the interventional Perturb-seq data to recover a superset of the surrogate parents. We first leverage the global Markov property and equivalently write: where denotes the surrogate ancestors of , or the variables in X associated with the ancestors of .

We discover the surrogate ancestors using unconditional independence tests. For any Xk ∈ X, we test by unpaired two-sided t-test, where Pi is an indicator function equal to one when we perturb Xi and zero in the control samples of Perturb-seq. Pi is thus a parent of Xi alone but not a child of B, so we do not need to condition on B. We use the two-sided t-test to assess for independence because the t-statistic averages over cells to mimic bulk RNA-seq. If we conclude that , then Xk must be a descendant of Pi by the global Markov property, so we include Xk into the set of surrogate descendants . Curating every Xj ∈ X such that into yields the surrogate ancestors of as desired.

Procedure

We now introduce an algorithm called Root Causal Strength using Perturbations (RCSP) that discovers the surrogate ancestors of each variable using Perturb-seq and then computes the RCS of each variable using bulk RNA-seq. We summarize RCSP in Algorithm 1.

RCSP takes Perturb-seq and bulk RNA-seq datasets as input. The algorithm first finds the surrogate descendants of each variable in in Line 2 in order to identify the surrogate ancestors of each variable in Line 5. Access to the surrogate ancestors and the batches B allows RCSP to compute Φi for each Xi ∈ X from the bulk RNA-seq in Line 6. The algorithm thus outputs the RCS scores Φ as desired.

Root Causal Strength using Perturbations (RCSP)

We certify RCSP as follows:

(Fisher consistency) Consider the same assumption as Lemma 1. If unconditional d-separation faithfulness holds, then RCSP recovers Φ almost surely as N → ∞.

We engineered RCSP to only require unconditional d-separation faithfulness because real distributions may not obey full d-separation faithfulness (Strobl, 2022).

Synthetic Data

Simulations

We generated a linear SEM obeying Equation (3) specifically as for every and similarly . We included p + 1 = 2500 variables in . We instantiated the coefficient matrix β by sampling from a Bernoulli(2/(p − 1)) distribution in the upper triangular portion of the matrix. The resultant causal graph thus has an expected neighborhood size of 2. We then randomly permuted the ordering of the variables. We introduced weights into the coefficient matrix by multiplying each entry in β by a weight sampled uniformly from [−1, −0.25] ∪ [0.25, 1]. The error terms each follow a standard Gaussian distribution multiplied by 0.5. We introduced batch effects by drawing each entry of the mapping efficiencies π from the uniform distribution between 10 and 1000 for the bulk RNA-seq, and between 0.1 and 1 for the Perturb-seq. We set and then obtained the corrupted surrogate Xi distributed for each and batch j. We chose Y uniformly at random from the set of vertices with at least one parent and no children. We repeated the above procedure 30 times.

Comparators

We compared RCSP against the following four algorithms:

Additive noise model (ANM) (Peters et al., 2014; Strobl and Lasko, 2023a): performs non-linear regression of Xi on Pa(Xi) ∪ B and then regresses Y on the residuals E \ Ei to estimate |𝔼(Y|E \ Ei) − 𝔼(Y |X, B) | for each Xi ∈ X. The non-linear regression residuals are equivalent to the error terms assuming an additive noise model.

Linear Non-Gaussian Acyclic Model (LiNGAM) (Peters et al., 2014; Strobl and Lasko, 2022): same as above but performs linear instead of non-linear regression.

CausalCell (Wen et al., 2023): selects the top 50 genes with maximal statistical dependence to Y, and then runs the Peter-Clark (PC) algorithm (Spirtes et al., 2000) using a non-parametric conditional independence test to identify a causal graph among the top 50 genes. The algorithm does not perform root causal inference, so we use ANM as above but condition on the estimated parent sets for the top 50 genes and the ancestors inferred from the Perturb-seq data otherwise.

Univariate regression residuals (Uni Reg): regresses Y on Xi ∪ B and estimates the absolute residuals |Y − 𝔼(Y|Xi, B) | for each Xi ∈ X.

Multivariate regression residuals (Multi Reg): similar to above but instead computes the absolute residuals after regressing Y on (X \ Xi) ∪ B.

The first two methods are state-of-the-art approaches used for root causal discovery. Univariate and multivariate regressions do not distinguish between predictivity and causality, but we included them as sanity checks. We performed all non-linear regressions using multivariate adaptive regression splines to control for the underlying regressor (Friedman, 1991). We compared the algorithms on their accuracy in estimating Φ.

Real Data

Quality Control

We downloaded Perturb-seq datasets of retinal pigment epithelial cells from the RPE-1 cell line, and lymphoblast cells from the K562 cell line (Replogle et al., 2022). We used the genome-wide dataset version for the latter. We downloaded the datasets from the scPerturb database on Zenodo (Green et al., 2022) with the same quality controls as the original paper. Replogle et al. computed adjusted library sizes by equalizing the mean library size of control cells within each batch. Cells with greater than a 2000 or 3000 library size, and less than 25% or 11% mitochondrial RNA were kept, respectively. The parameters were chosen by plotting the adjusted library sizes against the mitochondrial RNA counts and then manually setting thresholds that removed low quality cells likely consisting of ambient mRNA transcripts arising from premature cell lysis or cell death.

We next downloaded bulk RNA-seq datasets derived from patients with age-related macular degeneration (AMD; GSE115828) and multiple sclerosis (MS; GSE137143) (Ratnapriya et al., 2019; Kim et al., 2021). We excluded 10 individuals from the AMD dataset including one with an RNA integrity number of 21.92, five missing an integrity number (all others had an integrity number of less than 10), and four without a Minnesota Grading System score. We kept all samples from the MS dataset derived from CD4+ T cells but filtered out genes with a mean of less than 5 counts as done in the original paper.

We finally kept genes that were present in both the AMD bulk dataset and the RPE-1 Perturb-seq dataset, yielding a final count of 513 bulk RNA-seq samples and 247,914 Perturb-seq samples across 2077 genes. We also kept genes that were present in both the MS bulk dataset and the K562 Perturb-seq dataset, yielding a final count of 137 bulk RNA-seq samples and 1,989,578 Perturb-seq samples across 6882 genes. We included age and sex as a biological variable as covariates for every patient in both datasets in subsequent analyses.

Evaluation Rationale

We do not have access to the ground truth values of Φ in real data. We instead evaluate the RCSP estimates of Φ using alternative sources of ground truth knowledge. We first assess the accuracy of RCS using the control variable age as follows:

Determine if the RCS values of age identify age as a root cause in diseases that progress over time.

Second, root causal genes should imply a more omnigenic than polygenic model because the effects of a few error terms distribute over many downstream genes. We verify omnigenecity as follows:

Determine if the distribution of D-RCS concentrates around zero more than the distribution of the Deviation of Statistical Dependence (D-SD) defined as for each gene where Ωi = |𝔼(Y |Xi, B) − 𝔼(Y |B)| and ωij its value for patient j.

Despite the sparsity/omnigenecity of root causal genes, we still expect the root causal genes to correspond to at least some known causes of disease:

Determine if genes with the top D-RCS scores correspond to genes known to cause the disease.

Next, the root causal genes initiate pathogenesis, and we often have knowledge of pathogenic pathways even though we may not know the exact gene expression cascade leading to disease. Intervening on root causal genes should also modulate patient symptoms. We thus further evaluate the accuracy of RCSP using pathway and drug enrichment analyses as follows:

Determine if the D-RCS scores identify known pathogenic pathways of disease in pathway enrichment analysis.

Determine if the D-RCS scores identify drugs that treat the disease.

Finally, complex diseases frequently involve multiple pathogenic pathways that differ between patients. Patients with the same complex disease also respond differently to treatment. We hence evaluate the precision of RCS as follows:

Determine if the patient-specific RCS scores identify subgroups of patients involving different but still known pathogenic pathways.

Determine if the patient-specific RCS scores identify subgroups of patients that respond differently to drug treatment.

In summary, we evaluate RCSP in real data based on its ability to (1) identify age as a known root cause, (2) suggest an omnigenic root causal model, (3) recover known causal genes, (4) find known pathogenic pathways, (5) find drugs that treat the disease, and (6,7) delineate patient subgroups.

Enrichment Analyses

Multivariate adaptive regression splines introduce sparsity, but enrichment analysis performs better with a dense input. We can estimate the conditional expectations of Φ using any general non-linear regression method, so we instead estimated the expectations using kernel ridge regression equipped with a radial basis function kernel (Shawe-Taylor and Cristianini, 2004). We then computed the D-RCS across all patients for each variable in X. We ran pathway enrichment analysis using the fast gene set enrichment analysis (FGSEA) algorithm (Sergushichev, 2016) with one hundred thousand simple permutations using the D-RCS scores and pathway information from the Reactome database (version 1.86.0) (Fabregat et al., 2017). We likewise performed drug set enrichment analysis with the Drug Signature database (version 1.0) (Yoo et al., 2015). We repeated the above procedures for the D-RCS of any cluster identified by hierarchical clustering via Ward’s method (Ward Jr, 1963).

eLife Assessment

This work provides a potentially valuable framework for understanding the primary causes of disease. However, the evidence supporting the utility of the approach is incomplete given the reliance on strong assumptions about the underlying causal mechanisms.

Reviewer #1 (Public review):

Summary:

This manuscript seeks to estimate the causal effect of genes on disease. To do so, they introduce a novel algorithm, termed the Root Causal Strength using Perturbations (RCSP) algorithm. RCSP uses perturb-seq to first estimate the gene regulatory network structure among genes, and then uses bulk RNA-seq with phenotype data on the samples to estimate causal effects of genes on the phenotype conditional on the learned network structure. The authors assess the performance of RCSP in comparison to other methods via simulation. Next, they apply RCSP to two real human datasets: 513 individuals age-related macular degeneration and 137 individuals with multiple sclerosis.

Strengths:

The authors tackle an important and ambitious problem - the identification of causal contributors to disease in the context of a causal inference framework. As the authors point out, observational RNA-seq data is insufficient for this kind of causal discovery, since it is very challenging to recover the true underlying graph from observational data; interventional data are needed. However, little perturb-seq data has been generated with annotated phenotype data, and much bulk RNA-seq data has already been generated, so it is useful to propose an algorithm to integrate the two as the authors have done.

The authors also offer substantial theoretical exposition for their work, bringing to bear both the literature on causal discovery as well as literature on the genetic architecture of complex traits.

Weaknesses:

The notion of a "root" causal gene - which the authors define based on a graph theoretic notion of topologically sorting graphs - requires a graph that is directed and acyclic. It is the latter that constitutes an important weakness here - it simply is a large simplification of human biology to draw out a DAG including hundreds of genes and a phenotype Y and to claim that the true graph contains no cycles. This is briefly touched upon the discussion, but given the fundamental nature of this choice - the manuscript should devote at least some of the main results to exploring the consequence of mischaracterizing true cyclic graphs as DAGs in this framework. For example - consider the authors' analysis of T cell infiltration in multiple sclerosis (MS). CD4+ effector T cells have the interesting property that they are stimulated by IL2 as a growth factor; yet IL2 also stimulates the activation of (suppressive) regulatory T cells. What does it mean to analyze CD4+ regulation in disease with a graph that does not consider IL2 (or other cytokine) mediated feedback loops/cycles?

I also encourage the authors to consider more carefully when graph structure learned from perturb-seq can be ported over to bulk RNA-seq. Consider again the MS CD4+ example - the authors first start with a large perturb-seq experiment (Replogle et al., 2022) performed in K562 cells. To what extent are K562 cells, which are derived from a leukemia cell line, suitable for learning the regulatory structure of CD4+ cells from individuals with an MS diagnosis? Presumably this structure is not exactly correct - to what extent is the RCSP algorithm sensitive to false edges in this graph? This leap - from cell line to primary human cells - is also not modeled in the simulation. Although challenging - it would be ideal for the RCSP to model or reflect the challenges in correctly identifying the regulatory structure.

It should also be noted that in most perturb-seq experiments, the entire genome is not perturbed, and frequently important TFs (that presumably are very far "upstream" and thus candidate "root" causal genes) are not expressed highly enough to be detected with scRNA-seq. In that context - perhaps slightly modifying the language regarding RCSP's capabilities might be helpful for the manuscript - perhaps it would be better to describe it has an algorithm for causal discovery among a set of genes that were perturbed and measured, rather than a truly complete search for causal factors. Perhaps more broadly - it would also benefit the manuscript to devote slightly more text to describing the kinds of scenarios where RCSP (and similar ideas) would be most appropriately applied - perhaps a well-powered, phenotype annotated perturb-seq dataset performed in a disease relevant primary cell.

Reviewer #2 (Public review):

Summary:

This paper presents a very interesting use of a causal graph framework to identify the "root genes" of a disease phenotype. Root genes are the genes that cause a cascade of events that ultimately leads to the disease phenotype, assuming the disease progression is linear.

Strengths:

- The methodology has a solid theoretical background.

- This is a novel use of the causal graph framework to infer root causes in a graph

Weaknesses:

(1) General Comments

First, I have some general comments. I would argue that the main premise of the study might be inaccurate or incomplete. There are three major attributes of real biological systems, which are not considered in this work.

One is that the process from health-to-disease is not linear most of the time with many checks along the way that aim to prevent the disease phenotype. This leads to a non-deterministic nature of the path from health-to-disease. In other words, with the same root gene perturbations, and depending on other factors outside of gene expression, someone may develop a phenotype in a year, another in 10 years and someone else never. Claiming that this information is included in the error terms might not be sufficient to address this issue. The authors should discuss this limitation.

Two, the paper assumes that the network connectivity will remain the same after perturbation. This is not always true due to backup mechanisms in the cells. For example, suppose that a cell wants to create product P and it can do it through two alternative paths:

Path #1: A -> B -> P Path #2: A -> C -> P

Now suppose that path #1 is more efficient, so when B can be produced, path #2 is inactive. Once the perturbation blocks element B from being produced, the graph connectivity changes by activation of path #2. I did not see the authors taking this into consideration, which seems to be a major limitation in using perturb-seq results to infer connectivities.

Three, there is substantial system heterogeneity that may cause the same phenotype. This goes beyond the authors claim that although the initial gene causes of a disease may differ from person to person, at some point they will all converge to changes in the same set of "root genes". This is not true for many diseases, which are defined based on symptoms and lab tests at the patient level. You may have two completely different molecular pathologies that lead to the development of the same symptoms and test results. Breast cancer with its subtypes is a prime example of that. In theory, this issue could be addressed if there is infinite sample size. However, this assumption is largely violated in all existing biological datasets.

All the above limit the usefulness of this method for most chronic diseases, although it might still lead to interesting discoveries in cancer (in which the association between genes' dysregulation and development of cancer is more direct and occurs in less amount of time).

With these in mind, the theoretical and algorithmic advances this paper offers are interesting. And the theoretical proofs are solid.

(2) Specific comments.

I am curious how the simulated data were generated and processed. Specifically, were the values of the synthetic variables Z-scored? If not, then I would expect that the variance of every variable will increase from the roots of the graph to the leaves. That will give an advantage in any algorithm aiming to identify causal relations based on error terms. For fairness and completeness, the authors should Z-score the values in the synthetic data and compare the results.

The algorithm seems to require both RNA-seq and Perturb-seq data (Algorithm 1, page 14). Can it function with RNA-seq data only? What will be different in this case?

(3) Additional comments:

Although the manuscript is generally written clearly, some parts are not clear and others have missing details that make the narrative difficult to follow up. Some specific examples:

- Synthetic data generation: how many different graphs (SEMs) did they start from? (30?) How many samples per graph? Did they test different sample sizes?

- The presentation of comparative results (Suppl fig 4 and 7) is not clear. No details are given on how these results were generated. (what does it mean "The first column denotes the standard deviation of the outputs for each algorithm"?) Why all other methods have higher SD differences than RCSP? Is it a matter of scaling? Shouldn't they have at least some values near zero since the authors "added the minimum value so that all histograms begin at zero"? also, why RCSP results are more like a negative binomial distribution and every other is kind of normal?

- What is the significance of genes changing expression "from left to right" in a UMAP plot? (eg Fig. 3h and 3g)

The authors somewhat overstate the novelty of their algorithm. Representation of GRNs as causal graphs dates back in 2000 with the work of Nir Friedman in yeast. Other methods were developed more recently that look on regulatory network changes at the single sample level which the authors do not seem to be aware (e.g., Ellington et al, NeurIPS 2023 workshop GenBio and Bushur et al, 2019, Bioinformatics are two such examples). The methods they mention are for single cell data and they are not designed to connect single sample-level changes to a person's phenotype. The RCS method needs to be put in the right background context in order to bring up what is really novel about it.

Reviewer #3 (Public review):

Summary:

The authors provide an interesting and novel approach, RCSP, to determining what they call the "root causal genes" for a disease, i.e. the most upstream, initial causes of disease. RCSP leverages perturbation (e.g. Perturb-seq) and observational RNA-seq data, the latter from patients. They show using both theory and simulations that if their assumptions hold then the method performs remarkably well, compared to both simple and available state-of-the-art baselines. Whether the required assumptions hold for real diseases is questionable. They show superficially reasonable results on AMD and MS.

Strengths:

The idea of integrating perturbation and observational RNA-seq dataset to better understand the causal basis of disease is powerful and timely. We are just beginning to see genome-wide perturbation assay, albeit in limited cell-types currently. For many diseases, research cohorts have at least bulk observational RNA-seq from a/the disease-relevant tissue(s). Given this, RCSP's strategy of learning the required causal structure from perturbations and applying this knowledge in the observational context is pragmatic and will likely become widely applicable as Perturb-seq data in more cell-types/contexts becomes available.

The causal inference reasoning is another strength. A more obvious approach would be to attempt to learn the causal network structure from the perturbation data and leverage this in the observational data. However, structure learning in high-dimensions is notoriously difficult, despite recent innovations such as differentiable approaches. The authors notice that to estimate the root causal effect for a gene X, one only needs access to a (superset of) the causal ancestors of X: much easier relationships to detect than the full network.

The applications are also reasonably well chosen, being some of the few cases where genome-scale perturb-seq is available in a roughly appropriate (see below) cell-type, and observational RNA-seq is available at a reasonable sample size.

Weaknesses:

Several assumptions of the method are problematic. The most concerning is that the observational expression changes are all causally upstream of disease. There is work using Mendelian randomization (MR) showing that the _opposite_ is more likely to be true: most differential expression in disease cohorts is a consequence rather than a cause of disease (https://www.nature.com/articles/s41467-021-25805-y). Indeed, the oxidative stress of AMD has known cellular responses including the upregulation of p53. The authors need to think carefully about how this impacts their framework. Can the theory say anything in this light? Simulations could also be designed to address robustness.

A closely related issue is the DAG assumption of no cycles. This assumption is brought to bear because it required for much classical causal machinery, but is unrealistic in biology where feedback is pervasive. How robust is RCSP to (mild) violations of this assumption? Simulations would be a straightforward way to address this.

The authors spend considerable effort arguing that technical sampling noise in X can effectively be ignored (at least in bulk). While the mathematical arguments here are reasonable, they miss the bigger picture point that the measured gene expression X can only ever be a noisy/biased proxy for the expression changes that caused disease: 1) Those events happened before the disease manifested, possibly early in development for some conditions like neurodevelopmental disorders. 2) bulk RNA-seq gives only an average across cell-types, whereas specific cell-types are likely "causal". 3) only a small sample, at a single time point, is typically available. Expression in other parts of the tissue and at different times will be variable.

My remaining concerns are more minor.

While there are connections to the omnigenic model, the latter is somewhat misrepresented. 1) The authors refer to the "core genes" of the omnigenic model as being at the end (longitudinally) of pathogenesis. The omnigenic model makes no statements about temporally ordering: in causal inference terminology the core genes are simply the direct cause of disease. 2) "Complex diseases often have an overwhelming number of causes, but the root causal genes may only represent a small subset implicating a more omnigenic than polygenic model" A key observation underlying the omnigenic model is that genetic heritability is spread throughout the genome (and somewhat concentrated near genes expressed in disease relevant cell types). This implies that (almost) all expressed genes, or their associated (e)SNPs, are "root causes".

The claim that root causal genes would be good therapeutic targets feels unfounded. If these are highly variable across individuals then the choice of treatment becomes challenging. By contrast the causal effects may converge on core genes before impacting disease, so that intervening on the core genes might be preferable. The jury is still out on these questions, so the claim should at least be made hypothetical.

The closest thing to a gold standard I believe we have for "root causal genes" is integration of molecular QTLs and GWAS, specifically coloc/MR. Here the "E" of RCSP are explicitly represented as SNPs. I don't know if there is good data for AMD but there certainly is for MS. The authors should assess the overlap with their results. Another orthogonal avenue would be to check whether the root causal genes change early in disease progression.

The available perturb-seq datasets have limitations beyond on the control of the authors. 1) The set of genes that are perturbed. The authors address this by simply sub-setting their analysis to the intersection of genes represented in the perturbation and observational data. However, this may mean that a true ancestor of X is not modeled/perturbed, limiting the formal claims that can be made. Additionally, some proportion of genes that are nominally perturbed show little to no actual perturbation effect (for example, due to poor guide RNA choice) which will also lead to missing ancestors.

The authors provide no mechanism for statistical inference/significance for their results at either the individual or aggregated level. While I am a proponent of using effect sizes more than p-values, there is still value in understanding how much signal is present relative to a reasonable null.

I agree with the authors that age coming out of a "root cause" is potentially encouraging. However, it is also quite different in nature to expression, including being "measured" exactly. Will RCSP be biased towards variables that have lower measurement error?

Finally, it's a stretch to call K562 cells "lymphoblasts". They are more myeloid than lymphoid.

Author response:

The following is the authors’ response to the original reviews.

Reviewer 1:

(1) The notion of a “root” causal gene - which the authors define based on a graph theoretic notion of topologically sorting graphs - requires a graph that is directed and acyclic. It is the latter that constitutes an important weakness here - it simply is a large simplification of human biology to draw out a DAG including hundreds of genes and a phenotype Y and to claim that the true graph contains no cycles.

We agree that real causal graphs in biology often contain cycles. We now include additional experimental results with cyclic directed graphs in the Supplementary Materials. RCSP outperformed the other algorithms even in this setting, but we caution the reader that the theoretical interpretation of the RCS score may not coincide with a root causal effect when cycles exist:

“We also evaluated the algorithms on directed graphs with cycles. We generated a linear SEM over ρ + 1 = 1000 variables in . We sampled the coefficient matrix β from a Bernoulli (1/(p − 1)) distribution but did not restrict the non-zero coefficients to the upper triangular portion of the matrix. We then proceeded to permute the variable ordering and weight each entry as in the Methods for the DAG. We repeated this procedure 30 times and report the results in Supplementary Figure 3.

RCSP again outperformed all other algorithms even in the cyclic case. The results suggest that conditioning on the surrogate ancestors also estimates the RCS well even in the cyclic case. However, we caution that an error term Ei can affect the ancestors of when cycles exist. As a result, the RCS may not isolate the causal effect of the error term and thus not truly coincide with the notion of a root causal effect in cyclic causal graphs.”

(2) I also encourage the authors to consider more carefully when graph structure learned from Perturb-seq can be ported over to bulk RNA-seq. Presumably this structure is not exactly correct - to what extent is the RCSP algorithm sensitive to false edges in this graph? This leap - from cell line to primary human cells - is also not modeled in the simulation. Although challenging - it would be ideal for the RCSP to model or reflect the challenges in correctly identifying the regulatory structure.

We now include additional experimental results, where we gradually increased the incongruence between the DAG modeling the Perturb-seq and the DAG modeling the bulk RNA-seq using a mixture of graphs. The performance of RCSP degraded gradually, rather than abruptly, with increasing incongruence. We therefore conclude that RCSP is robust to differences between the causal graphs representing Perturb-seq and bulk RNA-seq:

“We next assessed the performance of RCSP when the DAG underlying the Perturb-seq data differs from the DAG underlying the bulk RNA-seq data. We considered a mixture of two random DAGs in bulk RNA-seq, where one of the DAGs coincided with the Perturb-seq DAG and second alternate DAG did not. We instantiated and simulated samples from each DAG as per the previous subsection. We generated 0%, 25%, 50%, 75%, and 100% of the bulk RNA-seq samples from the alternate DAG, and the rest from the Perturb-seq DAG. We ideally would like to see the performance of RCSP degrade gracefully, as opposed to abruptly, as the percent of samples derived from the alternate DAG increases.

We summarize results in Supplementary Figure 4. As expected, RCSP performed the best when we drew all samples from the same underlying DAG for Perturb-seq and bulk RNA-seq. However, the performance of RCSP also degraded slowly as the percent of samples increased from the alternate DAG. We conclude that RCSP can accommodate some differences between the underlying DAGs in Perturb-seq and bulk RNA-seq with only a mild degradation in performance.”

(3) It should also be noted that in most Perturb-seq experiments, the entire genome is not perturbed, and frequently important TFs (that presumably are very far “upstream” and thus candidate “root” causal genes) are not expressed highly enough to be detected with scRNA-seq. In that context - perhaps slightly modifying the language regarding RCSP’s capabilities might be helpful for the manuscript - perhaps it would be better to describe it as an algorithm for causal discovery among a set of genes that were perturbed and measured, rather than a truly complete search for causal factors. Perhaps more broadly it would also benefit the manuscript to devote slightly more text to describing the kinds of scenarios where RCSP (and similar ideas) would be most appropriately applied - perhaps a well-powered, phenotype annotated Perturb-seq dataset performed in a disease relevant primary cell.

We now clarify that Perturb-seq can only identify root causal genes among the perturbed set of genes in the Discussion:

“Modern genome-wide Perturb-seq datasets also adequately perturb and measure only a few thousand, rather than all, gene expression levels. RCSP can only identify root causal genes within this perturbed and measured subset.”

We now also describe the scenario where RCSP can identify root causal genes well in the Introduction:

“Experiments demonstrate marked improvements in performance, when investigators have access to a large bulk RNA-seq dataset and a genome-wide Perturb-seq dataset from a cell line of a disease-relevant tissue.”

Reviewer 2:

(1) The process from health-to-disease is not linear most of the time with many checks along the way that aim to prevent the disease phenotype. This leads to a non-deterministic nature of the path from health-to-disease. In other words, with the same root gene perturbations, and depending on other factors outside of gene expression, someone may develop a phenotype in a year, another in 10 years and someone else never. Claiming that this information is included in the error terms might not be sufficient to address this issue. The authors should discuss this limitation.

The proposed approach accommodates the above non-deterministic nature. The error terms of model factors that are outside of gene expression. We model the relation from gene expression to Y as probabilistic rather than deterministic because , where EY introduces stochasticity. Thus, two individuals with the same instantiations of the root causes may develop disease differently. We now clarify this in Methods:

“The error terms model root causes that are outside of gene expression, such as genetic variation or environmental factors. Moreover, the relation from gene expression to Y is stochastic because , where EY introduces the stochasticity. Two individuals may therefore have the exact same error term values over but different instantiations of Y.”

(2) The paper assumes that the network connectivity will remain the same after perturbation. This is not always true due to backup mechanisms in the cells. For example, suppose that a cell wants to create product P and it can do it through two alternative paths: Path #1: A → B → P, Path #2: A → C → P. Now suppose that path #1 is more efficient, so when B can be produced, path #2 is inactive. Once the perturbation blocks element B from being produced, the graph connectivity changes by activation of path #2. I did not see the authors taking this into consideration, which seems to be a major limitation in using Perturb-seq results to infer conductivities.

We agree that backup mechanisms can exist and therefore now include additional experimental results, where we gradually increased the incongruence between the DAG modeling the Perturb-seq and the DAG modeling the bulk RNA-seq using a mixture of graphs. The performance of RCSP degraded gradually, rather than abruptly, with increasing incongruence. We therefore conclude that RCSP is robust to differences between the causal graphs representing Perturb-seq and bulk RNA-seq:

“We next assessed the performance of RCSP when the DAG underlying the Perturb-seq data differs from the DAG underlying the bulk RNA-seq data. We considered a mixture of two random DAGs in bulk RNA-seq, where one of the DAGs coincided with the Perturb-seq DAG and second alternate DAG did not. We generated 0%, 25%, 50%, 75%, and 100% of the bulk RNA-seq samples from the alternate DAG, and the rest from the Perturb-seq DAG. We ideally would like to see the performance of RCSP degrade gracefully, as opposed to abruptly, as the percent of samples derived from the alternate DAG increases.

We summarize results in Supplementary Figure 4. As expected, RCSP performed the best when we drew all samples from the same underlying DAG for Perturb-seq and bulk RNA-seq. However, the performance of RCSP also degraded slowly as the percent of samples increased from the alternate DAG. We conclude that RCSP can accommodate some differences between the underlying DAGs in Perturb-seq and bulk RNA-seq with only a mild degradation in performance.”

(3) There is substantial system heterogeneity that may cause the same phenotype. This goes beyond the authors claim that although the initial gene causes of a disease may differ from person to person, at some point they will all converge to changes in the same set of “root genes.” This is not true for many diseases, which are defined based on symptoms and lab tests at the patient level. You may have two completely different molecular pathologies that lead to the development of the same symptoms and test results. Breast cancer with its subtypes is a prime example of that. In theory, this issue could be addressed if there is infinite sample size. However, this assumption is largely violated in all existing biological datasets.

The proposed method accommodates the above heterogeneity. We do not assume that the root causes affect the same set of root causal genes. Instead the root causes and root causal genes may vary from person to person. We write in the Introduction:

“The problem is further complicated by the existence of complex disease, where a patient may have multiple root causal genes that differ from other patients even within the same diagnostic category... We thus also seek to identify patient-specific root causal genes in order to classify patients into meaningful biological subgroups each hopefully dictated by only a small group of genes.”

The root causal genes may further affect different downstream genes at the patient-specific level. However root causal genes tend to have many downstream effects so that virtually every gene expression level becomes correlated with Y. We now clarify this by describing the omnigenic root causal model in the Introduction as follows:

“Finally, application of the algorithm to two complex diseases with disparate pathogeneses recovers an omnigenic root causal model, where a small set of root causal genes drive pathogenesis but impact many downstream genes within each patient. As a result, nearly all gene expression levels are correlated with the diagnosis at the population level.”

(4) Were the values of the synthetic variables Z-scored?

Yes, all variables were z-scored. We now clarify this in Methods:

“We also standardized all variables before running the regressions to prevent gaming of the marginal variances in causal discovery (Reisach et al., 2021; Ng et al., 2024).”

(5) The algorithm seems to require both RNA-seq and Perturb-seq data (Algorithm 1, page 14). Can it function with RNA-seq data only? What will be different in this case?

The algorithm cannot function with observational bulk RNA-seq data only. We included Perturb-seq because causal discovery with observational RNA-seq data alone tends to be inaccurate and unstable, as highlighted by the results of CausalCell. We further emphasize that we do not rely on d-separation faithfulness in Methods, which is typically required for causal discovery from observational data alone:

“We can also claim the backward direction under d-separation faithfulness. We however avoid making this additional assumption because real biological data may not arise from distributions obeying d-separation faithfulness in practice.”

(6) Synthetic data generation: how many different graphs (SEMs) did they start from? (30?) How many samples per graph? Did they test different sample sizes?

We now clarify that we generate 30 random SEMs, each associated with a DAG. We used 200 samples for the bulk RNA-seq to mimic a relatively large but common sample size. We also drew 200 samples for each perturbation or control in the Perturb-seq data. We did not consider multiple sample sizes due to the time required to complete each run. Instead, we focused on a typical scenario where investigators would apply RCSP. We now write the following in the Methods:

“We drew 200 samples for the bulk RNA-seq data to mimic a large but common dataset size. We introduced knockdown perturbations in Perturb-seq by subtracting an offset of two in the softplus function: . We finally drew 200 samples for the control and each perturbation condition to generate the Perturb-seq data. We repeated the above procedure 30 times.” We also include the following in Results:

“We obtained 200 cell samples from each perturbation, and another 200 controls without perturbations. We therefore generated a total of 2501 × 200 = 500,200 single cell samples for each Perturb-seq dataset. We simulated 200 bulk RNA-seq samples.”

(7) The presentation of comparative results (Supplementary Figures 4 and 7) is not clear. No details are given on how these results were generated. (what does it mean “The first column denotes the standard deviation of the outputs for each algorithm?”) Why all other methods have higher SD differences than RCSP? Is it a matter of scaling? Shouldn’t they have at least some values near zero since the authors “added the minimum value so that all histograms begin at zero?”

Each of these supplementary figures contains a 6 by 3 table of figures. By the first column, we mean column one (with rows 1 through 6) of each figure. The D-RCS and D-SD scores represent standard deviations of the RCS and SD scores from zero of each gene, respectively. We can similarly compute the standard deviation of the outputs of the algorithms. We now clarify this in the Supplementary Materials:

“The figure contains 6 rows and 3 columns. Similar to the D-RCS, we can compute the standard deviation of the output of each algorithm from zero for each gene. The first column in Supplementary Figure 7 denotes the histograms of these standard deviations across the genes.”

Many histograms do not appear to start at zero because the bars are too small to be visible. We now clarify this in the Supplementary Materials as well:

“Note that the bars at zero are not visible for many algorithms, since only a few genes attained standard deviations near the minimum.”

(8) Why RCSP results are more like a negative binomial distribution and every other is kind of normal?

All other methods have higher standard deviations than RCSP because they fail to compute an accurate measure of the root causal effect. Recall that, just like a machine has a few root causal problems, only a few root casual genes have large root causal effects under the omnigenic root causal model. The results of RCSP look more like a negative binomial distribution because most RCS scores are concentrated around zero and only a few RCS scores are large – consistent with the omnigenic root causal model. The other algorithms fail to properly control for the upstream genes and thus attain large standard deviations for nearly all genes. We now clarify these points in the Supplementary Materials as follows:

“If an algorithm accurately identifies root causal genes, then it should only identify a few genes with large conditional root causal effects under the omnigenic root causal model. The RCSP algorithm had a histogram with large probability mass centered around zero with a long tail to the right. The standard deviations of the outputs of the other algorithms attained large values for nearly all genes. Incorporating feature selection and causal discovery with CausalCell introduced more outliers in the histogram of ANM. We conclude that only RCSP detected an omnigenic root causal model.”

(9) What is the significance of genes changing expression “from left to right” in a UMAP plot? (e.g., Fig. 3h and 3g)

The first UMAP dimension captured the variability of the RCS scores for most root causal genes. As a result, we could focus our analysis on the black cluster in Figure 3 (g) with large RCS scores in the subsequent pathway enrichment analysis summarized in Figure 3 (j). If two dimensions were involved, then we would need to analyze at least two clusters (e.g., black and pink), but this was not the case. We now clarify this in Results:

“The RCS scores of most of the top genes exhibited a clear gradation increasing only from the left to the right hand side of the UMAP embedding; we plot an example in Figure 3 (h). We found three exceptions to this rule among the top 30 genes (example in Figure 3 (i) and see Supplementary Materials). RCSP thus detected genes with large RCS scores primarily in the black cluster of Figure 3 (g). Pathway enrichment analysis within this cluster alone yielded supra-significant results on the same pathway detected in the global analysis...”

(10) The authors somewhat overstate the novelty of their algorithm. Representation of GRNs as causal graphs dates back in 2000 with the work of Nir Friedman in yeast. Other methods were developed more recently that look on regulatory network changes at the single sample level which the authors do not seem to be aware (e.g., Ellington et al, NeurIPS 2023 workshop GenBio and Bushur et al, 2019, Bioinformatics are two such examples). The methods they mention are for single cell data and they are not designed to connect single sample-level changes to a person’s phenotype. The RCS method needs to be put in the right background context in order to bring up what is really novel about it.

We agree that many methods already exist for uncovering associational, predictive (Markov, neighborhood) and causal gene regulatory networks. We now cite the above papers. However, the novelty in our manuscript is not causal graph discovery, but rather estimation of root causal effects, detection of root causal genes, and the proposal of the omnigenic root causal model. We now clarify this in the

Introduction:

“Many algorithms focus on discovering associational or predictive relations, sometimes visually represented as gene regulatory networks (Costa et al., 2017; Ellington et al., 2023). Other methods even identify causal relations (Friedman et al., 2000; Wang et al., 2023; Wen et al., 2000; Buschur et al., 2000), but none pinpoint the first gene expression levels that ultimately generate the vast majority of pathogenesis. Simply learning a causal graph does not resolve the issue because causal graphs do not summarize the effects of unobserved root causes, such as unmeasured environmental changes or variants, that are needed to identify all root causal genes. We therefore define the Root Causal Strength (RCS) score...”

Reviewer 3:

(1) Several assumptions of the method are problematic. The most concerning is that the observational expression changes are all causally upstream of disease. There is work using Mendelian randomization (MR) showing that the opposite is more likely to be true: most differential expression in disease cohorts is a consequence rather than a cause of disease (Porcu et al., 2021). Indeed, the oxidative stress of AMD has known cellular responses including the upregulation of p53. The authors need to think carefully about how this impacts their framework. Can the theory say anything in this light? Simulations could also be designed to address robustness.

Strictly speaking, we believe that differential expression in disease most likely has a cyclic causal structure: gene expression causes a diagnosis or symptom severity, and a diagnosis or symptom severity lead to treatments and other behavioral changes that perturb gene expression. For example, revTMWR in Porcu et al. (2021) uses trans-variants that are less likely to directly cause gene expression and instead directly cause a phenotype. However, TWMR as proposed in Porcu et al. (2019) instead uses cis-eQTLs and finds many putative causal relations from gene expression to phenotype. Thus, both causal directions likely hold.

RCSP uses disease-relevant tissue believed to harbor gene expression levels that cause disease. However, RCSP theoretically cannot handle the scenario where Y is a non-sink vertex and is a parent of a gene expression level because modern Perturb-seq datasets usually do not perturb or measure Y. We therefore empirically investigated the degree of error by running experiments, where we set Y to a non-sink vertex, so that it can cause gene expression. We find that the performance of RCSP degrades considerably for gene expression levels that contain Y as a parent. Thus RCSP is sensitive to violations of the sink target assumption:

“We finally considered the scenario where Y is a non-sink (or non-terminal) vertex. If Y is a parent of a gene expression level, then we cannot properly condition on the parents because modern Perturbseq datasets usually do not intervene on Y or measure Y . We therefore empirically investigated the degradation in performance resulting from a non-sink target Y, in particular for gene expression levels where Y is a parent. We again simulated 200 samples from bulk RNA-seq and each condition of Perturbseq with a DAG over 1000 vertices, an expected neighborhood size of 2 and a non-sink target Y . We then removed the outgoing edges from Y and resampled the DAG with a sink target. We compare the results of RCSP for both DAGs in gene expression levels where Y is a parent. We plot the results in Supplementary Figure 5. As expected, we observe a degradation in performance when Y is not terminal, where the mean RMSE increased from 0.045 to 0.342. We conclude that RCSP is sensitive to violations of the sink target assumption.”

(2) A closely related issue is the DAG assumption of no cycles. This assumption is brought to bear because it is required for much classical causal machinery, but is unrealistic in biology where feedback is pervasive. How robust is RCSP to (mild) violations of this assumption? Simulations would be a straightforward way to address this.

We agree that real causal graphs in biology often contain cycles. We now include additional experimental results with cyclic directed graphs in the Supplementary Materials. RCSP outperformed the other algorithms even in this setting, but we caution the reader that the theoretical interpretation of the RCS score may not coincide with a root causal effect when cycles exist:

“We also evaluated the algorithms on directed graphs with cycles. We generated a linear SEM over p + 1 = 1000 variables in . We sampled the coefficient matrix β from a Bernoulli (1/(p − 1)) distribution but did not restrict the non-zero coefficients to the upper triangular portion of the matrix. We then proceeded to permute the variable ordering and weight each entry as in the Methods for the DAG. We repeated this procedure 30 times and report the results in Supplementary Figure 3.

RCSP again outperformed all other algorithms even in the cyclic case. The results suggest that conditioning on the surrogate ancestors also estimates the RCS well even in the cyclic case. However, we caution that an error term Ei can affect the ancestors of , when cycles exist. As a result, the RCS may not isolate the causal effect of the error term and thus not truly coincide with the notion of a root causal effect in cyclic causal graphs.”

(3) The authors spend considerable effort arguing that technical sampling noise in X can effectively be ignored (at least in bulk). While the mathematical arguments here are reasonable, they miss the bigger picture point that the measured gene expression X can only ever be a noisy/biased proxy for the expression changes that caused disease: 1) Those events happened before the disease manifested, possibly early in development for some conditions like neurodevelopmental disorders. 2) bulk RNA-seq gives only an average across cell-types, whereas specific cell-types are likely “causal.” 3) only a small sample, at a single time point, is typically available. Expression in other parts of the tissue and at different times will be variable.

We agree that many other sources of error exist. The causal model of RNA-expression in Methods corresponds to a single snapshot in time for each sample. We now clarify this in the Methods as follows:

“We represent a snapshot of a biological causal process using an SEM over obeying Equation (3).”

We thus only detect the root causal genes in a single snapshot in time for each sample in bulk RNA-seq. If we cannot detect the root causal effect in a gene due to the signal washing out over time as in (1), or if the root causal effect in different cell types cancel each other out to exactly zero in bulk as in (2), then we cannot detect those root causal genes even with an infinite sample size.

(4) While there are connections to the omnigenic model, the latter is somewhat misrepresented. The authors refer to the “core genes” of the omnigenic model as being at the end (longitudinal) of pathogenesis. The omnigenic model makes no statements about temporal ordering: in causal inference terminology the core genes are simply the direct causes of disease.

We now clarify that we use the word pathogenesis to mean the causal cascade from root causes to the diagnosis. In this case, the direct causes of the diagnosis correspond to the end of pathogenesis, while the root causes correspond to the beginning. For example, if , with Y a diagnosis, then X1 is a root causal gene while X2 is a core (direct causal) gene. We now clarify this in the Introduction:

“Root causes of disease correspond to the most upstream causes of a diagnosis with strong causal effects on the diagnosis. Pathogenesis refers to the causal cascade from root causes to the diagnosis. Genetic and non-genetic factors may act as root causes and affect gene expression as an intermediate step during pathogenesis. We introduce root causal gene expression levels – or root causal genes for short – that correspond to the initial changes to gene expression induced by genetic and non-genetic root causes that have large causal effects on a downstream diagnosis (Figure 1 (a)). Root causal genes differ from core genes that directly cause the diagnosis and thus lie at the end, rather than at the beginning, of pathogenesis (Boyle et al., 2017).”

(5) A key observation underlying the omnigenic model is that genetic heritability is spread throughout the genome (and somewhat concentrated near genes expressed in disease relevant cell types). This implies that (almost) all expressed genes, or their associated (e)SNPs, are “root causes”.

We now clarify that genetic heritability can be spread throughout the genome in the omnigenic root causal model as well in the Discussion:

“Further, each causal genetic variant tends to have only a small effect on disease risk in complex disease because the variant can directly cause Y or directly cause any causal gene including those with small root causal effects on Y ; thus, all error terms that cause Y can model genetic effects on Y. However, the root causal model further elaborates that genetic and non-genetic factors often combine to produce a few root causal genes with large root causal effects, where non-genetic factors typically account for the majority of the large effects in complex disease. Many variants may therefore cause many genes in diseases with only a few root causal genes.”

We finally add Figure 5 into the Discussion as a concrete example illustrating the omnigenic root causal model:

(6) The claim that root causal genes would be good therapeutic targets feels unfounded. If these are highly variable across individuals then the choice of treatment becomes challenging. By contrast the causal effects may converge on core genes before impacting disease, so that intervening on the core genes might be preferable. The jury is still out on these questions, so the claim should at least be made hypothetical.

We clarify that we do not claim that root causal genes are better treatment targets than core genes in terms of magnitudes of causal effects on the phenotype. For example, in the common cold with a virus as the root cause, giving a patient an antiviral will eliminate fever and congestion, but so will giving a decongestant and an antipyretic. We only claim that treating root causal genes can eliminate disease near its pathogenic onset, just like giving an antiviral can eliminate the viral load and stop pathogenesis. We write the following the Introduction:

“Treating root causal genes can modify disease pathogenesis in its entirety, whereas targeting other causes may only provide symptomatic relief... Identifying root causal genes is therefore critical for developing treatments that eliminate disease near its pathogenic onset.”

We also further clarify in the Discussion that root causal genes account for deleterious causal effects not captured by the diagnosis Y:

“We finally emphasize that the root causal model accounts for all deleterious effects of the root causal genes, whereas the core gene model only captures the deleterious effects captured by the diagnosis Y. For example, the disease of diabetes causes retinopathy, but retinopathy is not a part of the diagnostic criteria of diabetes. As a result, the gene expression levels that cause retinopathy but not the diagnosis of diabetes are not core genes, even though they are affected by the root causal genes.”

We do agree that root causal genes may differ substantially between patients, although it is unclear if the heterogeneity is too great to develop treatments.

(7) The closest thing to a gold standard I believe we have for “root causal genes” is integration of molecular QTLs and GWAS, specifically coloc/MR. Here the “E” of RCSP are explicitly represented as SNPs. I don’t know if there is good data for AMD but there certainly is for MS. The authors should assess the overlap with their results. Another orthogonal avenue would be to check whether the root causal genes change early in disease progression.

Colocalization and Mendelian randomization unfortunately cannot identify root causal effects because they all attempt, either heuristically (colocalization) or rigorously (MR), to identify variants that cause each gene expression level rather than variants that directly cause each gene expression level and thus make up the error terms. We therefore need new methods that can identify direct causal variants in order to assess overlap.

We checked whether root causal genes change early in disease progression using knowledge of pathogenesis. In particular, oxidative stress induces pathogenesis in AMD, and RCSP identified root causal genes involved in oxidative stress in AMD:

“The pathogenesis of AMD involves the loss of RPE cells. The RPE absorbs light in the back of the retina, but the combination of light and oxygen induces oxidative stress, and then a cascade of events such as immune cell activation, cellular senescence, drusen accumulation, neovascularization and ultimately fibrosis (Barouch et al., 2007). We therefore expect the root causal genes of AMD to include genes involved in oxidative stress during early pathogenesis. The gene MIPEP with the highest D-RCS score in Figure 3 (d) indeed promotes the maturation of oxidative phosphorylation-related proteins (Shi et al., 2011). The second gene SLC7A5 is a solute carrier that activates mTORC1 whose hyperactivation increases oxidative stress via lipid peroxidation (Nachef et al., 2021; Go et al., 2020). The gene HEATR1 is involved in ribosome biogenesis that is downregulated by oxidative stress (Turi et al., 2018). The top genes discovered by RCSP thus identify pathways known to be involved in oxidative stress.”

Similarly, T cell infiltration across the blood brain barrier initiates pathogenesis in MS, and RCSP identified root causal genes involved in this infiltration:

“Genes with the highest D-RCS scores included MNT, CERCAM and HERPUD2 (Figure 4 (d)). MNT is a MYC antagonist that modulates the proliferative and pro-survival signals of T cells after engagement of the T cell receptor (Gnanaprakasam et al., 2017). Similarly, CERCAM is an adhesion molecule expressed at high levels in microvessels of the brain that increases leukocyte transmigration across the blood brain barrier (Starzyk et al., 2000). HERPUD2 is involved in the endoplasmic-reticulum associated degradation of unfolded proteins (Kokame et al., 2000). Genes with the highest D-RCS scores thus serve key roles in known pathogenic pathways of MS.”

(8) The available Perturb-seq datasets have limitations beyond on the control of the authors. 1) The set of genes that are perturbed. The authors address this by simply sub-setting their analysis to the intersection of genes represented in the perturbation and observational data. However, this may mean that a true ancestor of X is not modeled/perturbed, limiting the formal claims that can be made. Additionally, some proportion of genes that are nominally perturbed show little to no actual perturbation effect (for example, due to poor guide RNA choice) which will also lead to missing ancestors.

We now clarify that Perturb-seq can only identify root causal genes among the adequately perturbed set of genes in the Discussion:

“Modern genome-wide Perturb-seq datasets also only adequately perturb and measure a few thousand, rather than all, gene expression levels. RCSP can only identify root causal genes within this perturbed and measured subset.”

(9) The authors provide no mechanism for statistical inference/significance for their results at either the individual or aggregated level. While I am a proponent of using effect sizes more than p-values, there is still value in understanding how much signal is present relative to a reasonable null.

We now explain that RCSP does not perform statistical inference in Methods because it is not clear how to define the appropriate cut-off for the RCS score under the null distribution:

“We focus on statistical estimation rather than statistical inference because Φi > 0 when Ei causes Y under mild conditions, so we reject the null hypothesis that Φi = 0 for many genes if many gene expression levels cause Y. However, just like a machine typically breaks down due to only one or a few root causal problems, we hypothesize that only a few genes have large RCS scores Φi ≫ 0 even in complex disease.”

(10) I agree with the authors that age coming out of a “root cause” is potentially encouraging. However, it is also quite different in nature to expression, including being “measured” exactly. Will RCSP be biased towards variables that have lower measurement error?

We tested the above hypothesis by plotting sequencing depth against the D-RCS scores of each gene. We observed a small negative correlation between sequencing depth and D-RCS scores, indicating the D-RCS scores are slightly biased upwards with low sequencing depth. However, genes with the largest D-RCS scores exhibited a wide variety of sequencing depths in both MS and AMD, suggesting that sequencing depth has minimal effect on the largest D-RCS scores. We now explain these results for AMD in the Supplementary Materials:

“Theorem 1 states that RCS scores may exhibit bias with insufficient sequencing depth. The genes with large D-RCS scores may therefore simply have low sequencing depths. To test this hypothesis, we plotted sequencing depth against D-RCS scores. Consistent with Theorem 1, we observed a small negative correlation between D-RCS and sequencing depth (ρ = −0.16, p=2.04E-13), and D-RCS scores exhibited greater variability at the lowest sequencing depths (Supplementary Figure 8). However, genes with the largest D-RCS scores had mean sequencing depths interspersed between 20 and 3000. We conclude that genes with the largest D-RCS scores had a variety of sequencing depths ranging from low to high.”

We also report the results for MS:

“We plot sequencing depth against the D-RCS scores of each gene similar to the AMD dataset. We again observed a small negative correlation (ρ = −0.136, p_<_2.2E-16), indicating that genes with low sequencing depths had slightly higher D-RCS scores on average (Supplementary Figure 12). However, genes with the largest D-RCS scores again had a variety of sequencing depths. We conclude that sequencing depth has minimal correlation with the largest D-RCS scores.”

(11) Finally, it’s a stretch to call K562 cells “lymphoblasts.” They are more myeloid than lymphoid.

We now clarify that K562 cells are undifferentiated blast cells that can be induced to differentiate into lymphoblasts in Results:

“We next ran RCSP on 137 samples collected from CD4+ T cells of multiple sclerosis (MS; GSE137143) as well as Perturb-seq data of 1,989,578 undifferentiated blast cells that can be induced to differentiate into lymphoblasts, or the precursors of T cells and other lymphocytes.”

Version 2

Abstract

Abstract

Root causal gene expression levels – or root causal genes for short – correspond to the initial changes to gene expression that generate patient symptoms as a downstream effect. Identifying root causal genes is critical towards developing treatments that modify disease near its onset, but no existing algorithms attempt to identify root causal genes from data. RNA-sequencing (RNA-seq) data introduces challenges such as measurement error, high dimensionality and non-linearity that compromise accurate estimation of root causal effects even with state-of-the-art approaches. We therefore instead leverage Perturb-seq, or high throughput perturbations with single cell RNA-seq readout, to learn the causal order between the genes. We then transfer the causal order to bulk RNA-seq and identify root causal genes specific to a given patient for the first time using a novel statistic. Experiments demonstrate large improvements in performance. Applications to macular degeneration and multiple sclerosis also reveal root causal genes that lie on known pathogenic pathways, delineate patient subgroups and implicate a newly defined omnigenic root causal model.

Introduction

Root causes of disease correspond to the most upstream causes of a diagnosis with strong causal effects on the diagnosis. Pathogenesis refers to the causal cascade from root causes to the diagnosis. Genetic and non-genetic factors may act as root causes and affect gene expression as an intermediate step during pathogenesis. We introduce root causal gene expression levels – or root causal genes for short – that correspond to the initial changes to gene expression induced by genetic and non-genetic root causes that have large causal effects on a downstream diagnosis (Figure 1 (a)). Root causal genes differ from core genes that directly cause the diagnosis and thus lie at the end, rather than at the beginning, of pathogenesis (Boyle et al., 2017). Root causal genes also generalize (the expression levels of) driver genes that only account for the effects of somatic mutations primarily in cancer (Martínez-Jiménez et al., 2020).

(a) Toy example where a variable E2 simultaneously models genetic and non-genetic root causes that jointly have a large causal effect on a diagnose Y through gene expression . E2 first affects the gene expression level , or the root causal gene. The root causal gene then affects other downstream levels during pathogenesis, including the core (or direct causal) gene , to ultimately induce a diagnosis Y. (b) We hypothesize that the causal effects of most root causes are small, but a few are large (red ellipse), in each patient with disease. As a result, the distribution of these root causal effects tends to be right skewed in disease.

Treating root causal genes can modify disease pathogenesis in its entirety, whereas targeting other causes may only provide symptomatic relief. For example, mutations in Gaucher disease cause decreased expression of wild type beta-glucocerebrosidase, or the root causal gene (Nagral, 2014). We can give a patient blood transfusions to alleviate the fatigue and anemia associated with the disease, but we seek more definitive treatments like recombinant glucocerebrosidase that replaces the deficient enzyme. Enzyme replacement therapy alleviates the associated liver, bone and neurological abnormalities of Gaucher disease as a downstream effect. Identifying root causal genes is therefore critical for developing treatments that eliminate disease near its pathogenic onset.

The problem is further complicated by the existence of complex disease, where a patient may have multiple root causal genes that differ from other patients even within the same diagnostic category (Cano-Gamez and Trynka, 2020). Complex diseases often have an overwhelming number of causes but, just like a machine usually breaks down due to one or a few root causal problems, the root causal genes may only represent a small subset of the genes because the causal effects of only a few root causes are large (Figure 1 (b)). We thus also seek to identify patient-specific root causal genes in order to classify patients into meaningful biological subgroups each hopefully dictated by only a small group of genes.

No existing method identifies root causal genes from data. Many algorithms focus on discov-ering associational or predictive relations, sometimes visually represented as gene regulatory net-works (Costa-Silva et al., 2017; Ellington et al., 2023). Other methods even identify causal relations (Friedman et al., 2000; Wang et al., 2023; Wen et al., 2023; Buschur et al., 2020), but none pinpoint the first gene expression levels that ultimately generate the vast majority of pathogen-esis. Simply learning a causal graph does not resolve the issue because causal graphs do not summarize the effects of unobserved root causes, such as unmeasured environmental changes or variants, that are needed to identify all root causal genes. We therefore define the Root Causal Strength (RCS) score to identify all root causal genes unique to each patient. We then design the Root Causal Strength using Perturbations (RCSP) algorithm that estimates RCS from bulk RNA-seq under minimal assumptions by integrating Perturb-seq, or high throughput perturbation experiments using CRISPR-based technologies coupled with single cell RNA-sequencing (Dixit et al., 2016; Adamson et al., 2016; Datlinger et al., 2017). Experiments demonstrate marked improvements in performance, when investigators have access to a large bulk RNA-seq dataset and a genomewide Perturb-seq dataset from a cell line of a disease-relevant tissue. Finally, application of the algorithm to two complex diseases with disparate pathogeneses recovers an omnigenic root causal model, where a small set of root causal genes drive pathogenesis but impact many downstream genes within each patient. As a result, nearly all gene expression levels are correlated with the diagnosis at the population level.

Results

We briefly summarize the Methods in the first two subsections.

Definitions

Differential expression analysis identifies differences in gene expression levels between groups Y (Costa-Silva et al., 2017). A gene Xi may be differentially expressed due to multiple reasons. For example, Xi may cause Y, or a confounder C may explain the relation between Xi and Y such that Xi ← C → Y. In this paper, we take expression analysis a step further by pinpointing causal relations from expression levels regardless of the variable type of Y (discrete or continuous). We in particular seek to discover patient-specific root causal genes from bulk RNA-seq data, which we carefully define below.

We represent a biological system in bulk RNA-seq as a causal graph 𝔾 – such as in Figure 2 (a) – where p vertices represent true gene expression levels in a bulk sample and Y denotes the patient symptoms or diagnosis. The set contains thousands of genes in practice. Directed edges between the vertices in 𝔾 refer to direct causal relations. We assume that gene expression causes patient symptoms but not vice versa so that no edge from Y is directed towards The set Pa refers to the parents of , or those variables with an edge directed into . For example, in Figure 2 (a). A root vertex corresponds to a vertex with no parents.

Method overview and synthetic data results. (a) We consider a latent causal graph over the true counts . (b) We augment the graph with error terms E such that each Ei ∈ E in red has an edge directed towards . (c) The RCS of , denoted by Φ2, quantifies the magnitude of the conditional root causal effect, or the strength of the causal effect from E2 to Y conditional on . (d) We cannot observe in practice but instead observe the noisy surrogates X in blue corrupted by Poisson measurement error. (e) Perturbing a variable such as changes the marginal distributions of downstream variables shown in green under mild conditions. (f) RCSP thus uses the perturbation data to identify (an appropriate superset of) the surrogate parents for each variable in order to compute Φ. (g) Violin plots show that RCSP achieved the smallest RMSE to the ground truth RCS values in the synthetic data. (h) RCSP also took about the same amount of time to complete as multivariate regression. Univariate regression only took 11 seconds on average, so its bar is not visible. Error bars denote 95% confidence intervals of the mean. (i) Finally, RCSP maintained low RMSE values regardless of the number of clusters considered.

We can associate 𝔾 with the structural equation for each that links each vertex to its parents and error term Ei (Pearl, 2009). The error term Ei is not simply a regression residual but instead represents the conglomeration of unobserved explanatory variables that only influence , such as unobserved transcriptional regulators, certain genetic variants and specific environmental conditions. We thus also include the error terms E in the directed graph of Figure 2 (b). All root vertices are error terms and vice versa. The root causes of Y are the error terms that cause Y, or have a directed path into Y. We define the root causal strength (RCS) of on Y as the following absolute difference (Figure 2 (c)): We prove the last equality in the Methods. As a result, RCS Φi directly measures the contribution of the gene on Y according to its error term Ei without recovering the error term values. The algorithm does not impose distributional assumptions or functional restrictions such as additive noise to estimate the error term values as an intermediate step. Moreover Φi is patient-specific because the values of and may differ between patients. We have Φi = 0 when Ei is not a cause of Y, and we say that the gene is a patient-specific root causal gene if Φi ≫ 0, or its (conditional) root causal effect is large as depicted by the red ellipse in Figure 1 (b).

Algorithm

We propose an algorithm called Root Causal Strength using Perturbations (RCSP) that estimates Φ = {Φ1, …, Φp} from genes measured in both bulk RNA-seq and Perturb-seq datasets derived from possibly independent studies but from the same tissue type. We rely on bulk RNA-seq instead of single cell RNA-seq in order to obtain many samples of the label Y. We focus on statistical estimation rather than statistical inference because Φi > 0 when Ei causes Y under mild conditions, so we reject the null hypothesis that Φi = 0 for many genes if many gene expression levels cause Y. However, just like a machine typically breaks down due to only one or a few root causal problems, we hypothesize that only a few genes have large RCS scores Φi ≫ 0 even in complex disease.

Estimating Φ requires access to the true gene expression levels and the removal of the effects of confounding. We first control for batch effects representing unwanted sources of technical variation such as different sequencing platforms or protocols. We however can only obtain imperfect counts X from RNA sequencing even within each batch (Figure 2 (d)). Measurement error introduces confounding as well because it prevents us from exactly controlling for the causal effects of the gene expression levels. Investigators usually mitigate measurement error by normalizing the gene expression levels by sequencing depth. We show in the Methods that the Poisson distribution approximates the measurement error distribution induced by the sequencing process to high accuracy (Choudhary and Satija, 2022; Sarkar and Stephens, 2021). We leverage this fact to eliminate the need for normalization by sequencing depth using an asymptotic argument where the library size N approaches infinity. N takes on a value of at least ten million in bulk RNA-seq, but we also empirically verify that the theoretical results hold well in the Supplementary Materials. We thus eliminate the Poisson measurement error and batch effects by controlling for the batches B but not N in non-linear regression models.

We in particular show that Φi in Equation (1) is also equivalent to: where refers to the surrogate parents of , or the variables in X associated with . RCSP can identify (an appropriate superset of) the surrogate parents of each variable using perturbation data because perturbing a gene changes the marginal distributions of its downstream effects – which the algorithm detects from data under mild assumptions (Figures 2 (e) and (f)). The algorithm thus only transfers the binary presence or absence of causal relations from the single cell to bulk data – rather than the exact functional relationships – in order to remain robust against discrepancies between the two data types; we empirically verify the robustness in the Supplementary Materials. RCSP finally performs the two non-linear regressions needed to estimate and for each Φi. We will compare Φi against Statistical Dependence (SD), a measure of correlational strength defined as Ωi = |𝔼(Y | Xi, B) − 𝔼(Y | B) | where we have removed the conditioning on .

In silico identification of root causal genes

We simulated 30 bulk RNA-seq and Perturb-seq datasets from random directed graphs summarizing causal relations between gene expression levels. We performed single gene knock-down perturbations over 2500 genes and 100 batches. We obtained 200 cell samples from each perturbation, and another 200 controls without perturbations. We therefore generated a total of 2501 × 200 = 500, 200 single cell samples for each Perturb-seq dataset. We simulated 200 bulk RNA-seq samples. We compared RCSP against the Additive Noise Model (ANM) (Peters et al., 2014; Strobl and Lasko, 2023a), the Linear Non-Gaussian Acyclic Model (LiNGAM) (Peters et al., 2014; Strobl and Lasko, 2022), CausalCell (Wen et al., 2023), univariate regression residuals (Uni Reg), and multivariate regression residuals (Multi Reg). The first two algorithms are state-of-the-art approaches used for error term extraction and, in theory, root causal discovery. See Methods for comprehensive descriptions of the simulation setup and comparator algorithms.

We summarize accuracy results in Figure 2 (g) using the Root Mean Squared Error (RMSE) to the ground truth Φ values. All statements about pairwise differences hold true at a Bonferonni corrected threshold of 0.05/5 according to paired two-sided t-tests, since we compared RCSP against a total of five algorithms. RCSP estimated Φ most accurately by a large margin. ANM and LiNGAM are theoretically correct under their respective assumptions, but they struggle to outperform standard multivariate regression due to the presence of measurement error in RNA-seq (Supplementary Materials). Feature selection and causal discovery with CausalCell did not improve performance. Univariate regression performed the worst, since it does not consider the interactions between variables. RCSP achieved the lowest RMSE while completing in about the same amount of time as multivariate regression on average (Figure 2 (h)). RCSP maintained the lowest RMSE even in the cyclic case, and the performance of the algorithm remained robust to differences between the directed graphs underlying the bulk RNA-seq and Perturb-seq data (Supplementary Materials). We conclude that RCSP both scalably and accurately estimates Φ.

We will cluster the RCS values in real data to find patient subgroups. We therefore also performed hierarchical clustering using Ward’s method (Ward Jr, 1963) on the values of Φ estimated by RCSP with the synthetic data. We then computed the RMSEs and averaged them within each cluster. We found that RCSP maintained low average RMSE values regardless of the number of clusters considered (Figure 2 (i)). We conclude that RCSP maintains accurate estimation of Φ across different numbers of clusters.

Oxidative stress in age-related macular degeneration

We ran RCSP on a bulk RNA-seq dataset of 513 individuals with age-related macular degeneration (AMD; GSE115828) and a Perturb-seq dataset of 247,914 cells generated from an immortalized retinal pigment epithelial (RPE) cell line (Ratnapriya et al., 2019; Replogle et al., 2022). The Perturbseq dataset contains knockdown experiments of 2,077 genes overlapping with the genes of the bulk dataset. We set the target Y to the Minnesota Grading System score, a measure of the severity of AMD based on stereoscopic color fundus photographs. We always included age and sex as a biological variable as covariates. We do not have access to the ground truth values of Φ in real data, so we evaluated RCSP using seven alternative techniques. See Methods for a detailed rationale of the evaluation of real data. RCSP outperformed all other algorithms in this dataset (Supplementary Materials). We therefore only analyze the output of RCSP in detail here.

AMD is a neurodegenerative disease of the aging retina (Hadziahmetovic and Malek, 2021), so age is a known root cause of the disease. We therefore determined if RCSP identified age as a root cause. Note that RCSP does not need perturbation data of age to compute the RCS values of age, since age has no parents in the directed graph. The algorithm estimated a heavy tailed distribution of the RCS values indicating that most of the RCS values deviated away from zero (Figure 3 (a)). The Deviation of the RCS (D-RCS), or the standard deviation from an RCS value of zero, measures the tailedness of the distribution while preserving the unit of measurement. The D-RCS of age corresponded to 0.46 – more than double that of the nearest gene (Figure 3 (d)). We conclude that RCSP correctly detected age as a root cause of AMD.

Analysis of AMD. (a) The distribution of the RCS scores of age deviated away from zero and had a composite D-RCS of 0.46. (b) However, the majority of gene D-RCS scores concentrated around zero, whereas the majority of gene D-SD scores concentrated around the relatively larger value of 0.10. Furthermore, the D-RCS scores of the genes in (d) mapped onto the “amino acid transport across the plasma membrane” pathway known to be involved in the pathogenesis of AMD in (c). Blue bars survived 5% FDR correction. (e) Drug enrichment analysis revealed four significant drugs, the later three of which have therapeutic potential. (f) Hierarchical clustering revealed four clear clusters according to the elbow method, which we plot by UMAP dimensionality reduction in (g). The RCS scores of the top genes in (d) increased only from the left to right on the first UMAP dimension (x-axis); we provide an example of SLC7A5 in (h) and one of three detected exceptions in (i). We therefore performed pathway enrichment analysis on the black cluster in (g) containing the largest RCS scores. (j) The amino acid transport pathway had a larger degree of enrichment in the black cluster as compared to the global analysis in (c).

Root causal genes typically affect many downstream genes before affecting Y. We therefore expect to identify few root causal genes but many genes that correlate with Y. To evaluate this hypothesis, we examined the distribution of D-RCS relative to the distribution of the Deviation of Statistical Dependence (D-SD), or the standard deviation from an SD value of zero, in Figure 3 (b). Notice that the histogram of D-RCS scores in Figure 3 (b) mimics a folded distribution of Figure 1 (b). Thus, few D-RCS scores had large values implying the existence of only a few root causal genes. In contrast, most of the D-SD scores had relatively larger values concentrated around 0.10 implying the existence of many genes correlated with Y. We conclude that RCSP identified few root causal genes rather than many correlated genes for AMD.

The pathogenesis of AMD involves the loss of RPE cells. The RPE absorbs light in the back of the retina, but the combination of light and oxygen induces oxidative stress, and then a cascade of events such as immune cell activation, cellular senescence, drusen accumulation, neovascularization and ultimately fibrosis (Barouch and Miller, 2007). We therefore expect the root causal genes of AMD to include genes involved in oxidative stress during early pathogenesis. The gene MIPEP with the highest D-RCS score in Figure 3 (d) indeed promotes the maturation of oxidative phosphorylation-related proteins (Shi et al., 2011). The second gene SLC7A5 is a solute carrier that activates mTORC1 whose hyperactivation increases oxidative stress via lipid peroxidation (Nachef et al., 2021; Go et al., 2020). The gene HEATR1 is involved in ribosome biogenesis that is downregulated by oxidative stress (Turi et al., 2018). The top genes discovered by RCSP thus identify pathways known to be involved in oxidative stress. We further verified that measurement error did not explain their large D-RCS scores in Supplementary Materials.

We subsequently jointly analyzed the D-RCS values of all 2077 genes. We performed pathway enrichment analysis that yielded one pathway “amino acid transport across the plasma membrane” that passed an FDR threshold of 5% (Figure 3 (c)). The leading edge genes of the pathway included the solute carriers SLC7A5 and SLC1A5. These two genes function in conjunction to increase the efflux of essential amino acids out of the lysosome (Nicklin et al., 2009; Beaumatin et al., 2019). Some of these essential amino acids like L-leucine and L-arginine activate mTORC1 that in turn increases lipid peroxidation induced oxidative stress and the subsequent degeneration of the RPE (Nachef et al., 2021; Go et al., 2020). We conclude that pathway enrichment analysis correctly identified solute carrier genes involved in a known pathway promoting oxidative stress in AMD.

We next ran drug enrichment analysis with the D-RCS scores. The top compound arsenous acid inhibits RPE proliferation (Su et al., 2020), but the other three significant drugs have therapeutic potential (Figure 3 (e)). Busulfan decreases the requirement for intravitreal anti-VEGF injections (Dalvin et al., 2022). Genistein is a protein kinase inhibitor that similarly attenuates neovascularization (Kinoshita et al., 2014) and blunts the effect of ischemia on the retina (Kamalden et al., 2011). Finally, a metabolite of the antiviral agent 3’-azido-3’-deoxythymidine inhibits neovascularization and mitigates RPE degeneration (Narendran et al., 2020). We conclude that the D-RCS scores identified promising drugs for the treatment of AMD.

Hierarchical clustering and UMAP dimensionality reduction on the patient-specific RCS values revealed four clear clusters of patients by the elbow method on the sum of squares plot (Figures 3 (f) and (g), respectively). The RCS scores of most of the top genes exhibited a clear gradation increasing only from the left to the right hand side of the UMAP embedding; we plot an example in Figure 3 (h). We found three exceptions to this rule among the top 30 genes (example in Figure 3 (i) and see Supplementary Materials). RCSP thus detected genes with large RCS scores primarily in the black cluster of Figure 3 (g). Pathway enrichment analysis within this cluster alone yielded supra-significant results on the same pathway detected in the global analysis (Figure 3 (j) versus Figure 3 (c)). Furthermore, drug enrichment analysis results by cluster confirmed that patients in the black cluster with many root causal genes are likely the hardest to treat (Supplementary Materials). We conclude that RCSP detected a subgroup of patients whose root causal genes have large RCS scores and involve known pathogenic pathways related to oxidative stress.

T cell infiltration in multiple sclerosis

We next ran RCSP on 137 samples collected from CD4+ T cells of multiple sclerosis (MS; GSE137143) as well as Perturb-seq data of 1,989,578 undifferentiated blast cells that can be induced to differentiate into lymphoblasts, or the precursors of T cells and other lymphocytes (Kim et al., 2021; Replogle et al., 2022). We set the target Y to the Expanded Disability Status Scale score, a measure of MS severity. RCSP outperformed all other algorithms in this dataset as well (Supplementary Materials).

MS progresses over time, and RCSP correctly detected age as a root cause of MS severity with RCS values deviating away from zero (Figure 4 (a)). The distribution of gene D-RCS scores concentrated around zero with a long tail, whereas the distribution of gene D-SD scores concentrated around a relatively larger value of 0.3 (Figure 4 (b)). RCSP thus detected an omnigenic root causal model with a few root causal genes but many correlated genes.

Analysis of MS. (a) The distribution of the RCS scores of age deviated away from zero with a composite D-RCS of 0.55. (b) The distribution of D-RCS concentrated around zero, whereas the distribution of D-SD concentrated around 0.3. (d) RCSP identified many genes with large D-RCS scores that in turn mapped onto known pathogenic pathways in MS in (c). Hierarchical clustering revealed three clusters in (e), which we plot in two dimensions with UMAP in (f). Top genes did not correlate with either dimension of the UMAP embedding; we provide an example of the MNT gene in (g). (h) Drug enrichment analysis in the green cluster implicated multiple cathepsin inhibitors. Finally, EPH-ephrin signaling survived FDR correction in (c) and was enriched in the pink cluster in (i) which contained more MS patients with the relapsing-remitting subtype in (j); subtypes include relapse-remitting (RR), primary progressive (PP), secondary progressive (SP), clinically isolated syndrome (CIS), and radiologically isolated syndrome (RIS).

MS is an inflammatory neurodegenerative disease that damages the myelin sheaths of nerve cells in the brain and spinal cord. T cells may mediate the inflammatory process by crossing a disrupted blood brain barrier and repeatedly attacking the myelin sheaths (Fletcher et al., 2010). Damage induced by the T cells also perturbs cellular homeostasis and leads to the accumulation of misfolded proteins (Andhavarapu et al., 2019). The root causal genes of MS thus likely include genes involved in T cell infiltration across the blood brain barrier.

Genes with the highest D-RCS scores included MNT, CERCAM and HERPUD2 (Figure 4 (d)). MNT is a MYC antagonist that modulates the proliferative and pro-survival signals of T cells after engagement of the T cell receptor (Gnanaprakasam and Wang, 2017). Similarly, CERCAM is an adhesion molecule expressed at high levels in microvessels of the brain that increases leukocyte transmigration across the blood brain barrier (Starzyk et al., 2000). HERPUD2 is involved in the endoplasmic-reticulum associated degradation of unfolded proteins (Kokame et al., 2000). Genes with the highest D-RCS scores thus serve key roles in known pathogenic pathways of MS.

We found multiple genes with high D-RCS scores in MS, in contrast to AMD where age dominated (Figure 4 (d) versus Figure 3 (d)). Measurement error did not account for the high scores (Supplementary Materials). We performed pathway enrichment analysis using the D-RCS scores of all genes and discovered two significant pathways at an FDR corrected threshold of 5%: “adenomatous polyposis coli (APC) truncation mutants have impaired AXIN binding” and “EPH-ephrin signaling” (Figure 4 (c)). APC and AXIN are both members of the Wnt signaling pathway and regulate levels of beta-catenin (Spink et al., 2000). Furthermore, inhibition of Wnt/beta-catenin causes CD4+ T cell infiltration into the central nervous system via the blood brain barrier in MS (Lengfeld et al., 2017). Ephrins similarly regulate T cell migration into the central nervous system (Luo et al., 2016) and are overexpressed in MS lesions (Sobel, 2005). The APC-AXIN and EPH-ephrin pathways are thus consistent with the known pathophysiology of central nervous system T cell infiltration in MS.

We subsequently performed hierarchical clustering of the RCS scores. The within cluster sum of squares plot in Figure 4 (e) revealed the presence of three clusters by the elbow method. We plot the three clusters in a UMAP embedding in Figure 4 (f). The clusters did not show a clear relationship with MS symptom severity (Supplementary Materials) or the levels of the top most genes of Figure 4 (d); we plot the MNT gene as an example in Figure 4 (g). However, further analyses with additional genes revealed that the distribution of many lower ranked genes governed the structure of the UMAP embedding (Supplementary Materials). The D-RCS scores of each cluster also implicated different mechanisms of T cell pathology including APC-AXIN in the green cluster, disturbed T cell homeostasis in the pink cluster and platelet enhanced T cell autoreactivity in the blue cluster (Supplementary Materials).

Global drug enrichment analysis did not yield any significant drugs even at a liberal FDR thresh-old of 10%. We thus ran drug enrichment analysis in each cluster of Figure 4 (f). The blue and pink clusters again did not yield significant drugs. However, the third green cluster identified the cysteine cathepsin inhibitors dipeptide-derived nitriles, phenylalinine derivatives, e-64, L-006235 and L-873724 (Figure 4 (h)); statistical significance of the first three held even after correcting for multiple comparisons with the Bonferroni adjustment of 0.05/4 on the q-values. The leading edge genes of the significant drugs included the cathepsins CTSL, CTSS and CTSB exclusively. These drug enrichment results corroborate multiple experimental findings highlighting the therapeutic efficacy of cathepsin inhibitors in a subgroup of MS patients responsive to interferon therapy (Haves-Zburof et al., 2011; Burster et al., 2007).

Prior research has also shown that EPH-ephrin signaling is more prevalent in relapsing-remitting multiple sclerosis than in other subtypes of the disease (Golan et al., 2021). EPH-ephrin signaling survived FDR correction in our analysis (Figure 4 (c)). Furthermore, the pathway was more enriched in the pink cluster than in the other two (Figure 4 (i)). The pink cluster indeed contained a higher proportion of patients with the relapsing-remitting subtype (Figure 4 (j)). RCSP thus precisely identified the enrichment of EPH-ephrin signaling in the correct subtype of MS.

Discussion

We presented a framework for identifying root causal genes, or the gene expression levels directly regulated by root causes with large causal effects on Y, by modeling the root causes using the error terms of structural equation models. Each error term represents the conglomeration of unobserved root causes, such as genetic variants or environmental conditions, that directly cause a specific gene. We however do not have access to many of the error terms in practice, so we introduced the root causal strength (RCS) score, or the magnitude of the conditional causal effect of each error term, which we can compute using gene expression levels alone. The RCSP algorithm computes RCS given knowledge of the causal ancestors of each variable, which we obtained by Perturb-seq. RCSP only transfers the causal structure (binary cause-effect relations) from the single cell to bulk data rather than the exact functional relationships in order to remain robust against discrepancies between the two data types. Results with the synthetic data demonstrated marked improvements over existing alternatives. The algorithm also recovered only a few root causal genes that play key roles in known pathogenic pathways and implicate therapeutic drugs in both AMD and MS.

We detected a modest number of root causal genes in both AMD and MS, but virtually all genes were correlated with Y. This omnigenic model, where “omni-” refers to the nearly all genes correlated with Y, differs from the omnigenic model involving core genes (Boyle et al., 2017). Boyle et al. define core genes as genes that directly affect disease risk. The authors further elaborate that many peripheral genes affect the functions of a modest number of core genes, so the peripheral genes often explain most of disease heritability. In contrast, root causal genes may not directly cause Y but lie substantially upstream of Y in the causal graph. The error terms of upstream root causal genes affect many downstream genes that include both ancestors and non-ancestors of Y (Figure 5). These downstream genes contain traces of the root causal gene error terms that induce the many correlations with Y. The root causal model thus assumes sparsity in upstream root causal genes, whereas the core gene model assumes sparsity in the downstream direct causal genes (The omnigenic root causal model makes no statement about the number of direct causal genes, so direct causal genes may be sparse or dense). Further, each causal genetic variant tends to have only a small effect on disease risk in complex disease because the variant can directly cause Y or directly cause any causal gene including those with small root causal effects on Y; thus, all error terms that cause Y can model genetic effects on Y. However, the root causal model further elaborates that genetic and non-genetic factors often combine to produce a few root causal genes with large root causal effects, where non-genetic factors typically account for the majority of the large effects in complex disease. Many variants may therefore cause many genes in diseases with only a few root causal genes. We finally emphasize that the root causal model accounts for all deleterious effects of the root causal genes, whereas the core gene model only captures the deleterious effects captured by the diagnosis Y. For example, the disease of diabetes causes retinopathy, but retinopathy is not a part of the diagnostic criteria of diabetes. As a result, the gene expression levels that cause retinopathy but not the diagnosis of diabetes are not core genes, even though they are affected by the root causal genes. The sparsity of the root causal genes, the focus on the combined effects of genetic and non-genetic root causes, and the ability to account for root causal effects not represented by the target Y motivate us to use the phrase omnigenic root causal model in order to distinguish it from the omnigenic core gene model.

In this example, two root causal genes and affect many downstream genes and ultimately cause Y. Thus all genes correlate with Y, but only and have large root causal effects on Y. The omnigenic root causal model posits that only a few root causal genes affect many downstream genes, so that nearly all genes are correlated with Y. Causal genetic variants can directly cause Y or cause any gene expression level that causes Y – including those with small root causal effects – but only and have large root causal effects on Y due to genetic and non-genetic root causes modeled by E1 and E2. In contrast, the core gene model assumes only a few direct causal genes . These core genes do not account for the deleterious causal effects of E1 and E2 on and .

We identified root causal genes without imposing parametric assumptions using the RCS metric. Prior measures of root causal effect require restrictive functional relations, such as linear relations or additive noise, and continuous random variables (Strobl and Lasko, 2022; Strobl et al., 2024; Strobl and Lasko, 2023a). These restrictions ensure exact identifiability of the underlying causal graph and error terms. However, real RNA-seq is obtained from a noisy sequencing process and contains count data arguably corrupted by Poisson measurement error (Sarkar and Stephens, 2021). The Poisson measurement error introduces confounding that precludes exact recovery of the underlying error terms. The one existing root causal discovery method that can handle Poisson measurement error uses single cell RNA-seq, estimates negative binomial distribution parameters and cannot scale to the thousands of genes required for meaningful root causal detection (Strobl and Lasko, 2023b). RCSP rectifies the deficiencies of these past approaches by ensuring accurate root causal detection even in the presence of the counts, measurement error and high dimensionality of RNA-seq.

This study carries other limitations worthy of addressing in future work. The RCS score importantly quantifies root causal strength rather than root causal effect. As a result, the method cannot be used to identify the direction of root causal effect unconditional on the parents. The root causal effect and signed RCS (or expected conditional root causal effect) do not differ by much in practice (Supplementary Materials), but future work may focus on exactly identifying both the strength and direction of the unconditional causal effects of the error terms. Furthermore, RCS achieves patient but not cell-specificity because the algorithm relies on phenotypic labels obtained from bulk RNAseq. RCSP thus cannot identify the potentially different root causal genes present within distinct cell populations. Modern genome-wide Perturb-seq datasets also adequately perturb and measure only a few thousand, rather than all, gene expression levels. RCSP can only identify root causal genes within this perturbed and measured subset. Fourth, RCSP accounts for known batch effects and measurement error but cannot adjust for unknown confounding. Finally, RCSP assumes a directed acyclic graph. We can transform a directed graph with cycles into an acyclic one under equilibrium, but real biological distributions vary across time (Spirtes, 1995; Bongers et al., 2021). Future work should thus aim to estimate cell-specific root causal effects under latent confounding and time-varying distributions.

In conclusion, RCSP integrates bulk RNA-seq and Perturb-seq to identify patient-specific root causal genes under a principled causal inference framework using the RCS score. RCS quantifies root causal strength implicitly without requiring normalization by sequencing depth or direct access to the error terms of a structural equation model. The algorithm identifies the necessary causal relations to compute RCS using reliable high throughput perturbation data rather than observational data alone. The RCS scores often suggest an omnigenic root causal model of disease. Enrichment analyses with the RCS scores frequently reveal pathogenic pathways and drug candidates. We conclude that RCSP is a novel, accurate, scalable and disease-agnostic procedure for performing patient-specific root causal gene discovery.

Methods and Materials

Background on Causal Discovery

We denote a singleton variable like with italics and sets of variables like with bold italics. We can represent a causal process using a structural equation model (SEM) linking the p + 1 variables in using a series of deterministic functions: where fi is a function of the parents, or direct causes, of Zi and an error term Ei ∈ E. The error terms E are mutually independent. We will use the terms vertex and variable interchangeably. A root vertex corresponds to a vertex without any parents. On the other hand, a terminal or sink vertex is not a parent of any other vertex.

We can associate a directed graph to Z by drawing a directed edge from each member of Pa(Zi) to Zi for all Zi ∈ Z. A directed path from Zi to Zj corresponds to a sequence of adjacent directed edges from Zi to Zj. If such a path exists (or Zi = Zj), then Zi is an ancestor of Zj and Zj is a descendant of Zi. We collate all ancestors of Zi into the set Anc(Zi). A cycle occurs when there exists a directed path from Zi to Zj and the directed edge Zj → Zi. A directed acyclic graph (DAG) contains no cycles. We augment a directed graph by including additional vertices E and drawing a directed edge from each Ei ∈ E to Xi except when Xi = Ei is already a root vertex. We consider an augmented DAG 𝔾 throughout the remainder of this manuscript.

The vertices Zi and Zj are d-connected given W ⊆ Z \{Zi, Zj} in 𝔾 if there exists a path between Zi and Zj such that every collider on the path is an ancestor of W and no non-collider is in W. The vertices are d-separated if they are not d-connected. Any DAG associated with the SEM in Equation (3) also obeys the global Markov property where Zi and Zj are conditionally independent given W if they are d-separated given W. The term d-separation faithfulness refers to the converse of the global Markov property where conditional independence implies d-separation. A distribution obeys unconditional d-separation faithfulness when we can only guarantee d-separation faithfulness when W = ∅.

Causal Modeling of RNA Sequencing

Performing causal discovery requires careful consideration of the underlying generative process. We therefore propose a causal model for RNA-seq. We differentiate between the biology and the RNA sequencing technology.

We represent a snapshot of a biological causal process using an SEM over obeying Equation (3). We assume that the phenotypic target Y is a terminal vertex so that gene expression causes phenotype but not vice versa. Each corresponds to the total number of RNA molecules of a unique gene in a single cell or bulk tissue sample. The error terms model root causes that are outside of gene expression, such as genetic variation or environmental factors. Moreover, the relation from gene expression to Y is stochastic because Y = fY (Pa(Y), EY), where EY introduces the stochasticity. Two individuals may therefore have the exact same error term values over but different instantiations of Y.

We unfortunately cannot observe in practice but instead measure a corrupted count X using single cell or bulk RNA-seq technology. We derive the measurement error distribution from first principles. We map an exceedingly small fraction of each within a sample at unequal coverage. Let πij denote the probability of mapping one molecule of in batch j so that is near zero. The law of rare events (Papoulis, 1984) implies that the Poisson distribution wellapproximates the library size N so that .

We write the probability of mapping in a given sample as: This proportion remains virtually unchanged when sampling without replacement because with small . We can therefore approximate sampling without replacement by sampling with replacement using a multinomial: X ∼ MN(N; P1j, …, Ppj). This multinomial and the Poisson distribution over N together imply that the marginal distribution of each Xi ∈ X follows an independent Poisson distribution centered at , or: We conclude that the measurement error distribution follows a Poisson distribution to high accuracy. Multiple experimental results already corroborate this theoretical conclusion (Grün et al., 2014; Sarkar and Stephens, 2021; Choudhary and Satija, 2022).

We can represent the biology and the RNA sequencing in a single DAG over , where B denotes the batch, and Y the target variable representing patient symptoms or diagnosis. We provide a toy example in Figure 6. We draw 𝔾 over Z in black and make each a parent of Xi ∈ X in blue. We then include the root vertex B as a parent of all members of X in green. We augment this graph with the error terms of in red and henceforth refer to the augmented DAG as 𝔾. Repeated draws from the represented causal process generates a dataset.

An example of a DAG over augmented with the error terms E. The observed vertices X denote counts corrupted by batch B effects and Poisson measurement error.

No Need for Normalization by Sequencing Depth

We provide an asymptotic argument that eliminates the need for normalization by sequencing depth when estimating conditional expectations using bulk RNA-seq. The argument applies to the conditional expectations as a whole rather than their individual parameters.

We want to recover the causal relations between by removing batch B and depth N effects from the dataset because they correspond to the sequencing process rather than the underlying biology. We first consider removing sequencing depth by finding stably expressed housekeeping genes. Let Ã denote the set of housekeeping genes where is a constant for each similarly A refers to the corresponding set with Poisson measurement error. Let N = n be large enough such that for each sample. Then dividing by controls for sequencing depth in the following sense: where we have divided by a constant in the last term. Thus, dividing by L removes measurement error within each batch as N → ∞. We assume that N is so large that the approximation error is negligible. We only invoke the assumption in bulk RNA-seq, where the library size N is on the order of at least tens of millions.

We do not divide by L in practice because we may have L = 0 with finite N. We instead always include L ∪ B in the predictor set of downstream regressions. Conditioning on L ∪ B ensures that all downstream regressions mitigate depth and batch effects with adequate sequencing depth, or that 𝔼(Y | Ũ, B) = 𝔼(Y | U, L, B) for any as N → ∞. The equality holds almost surely under a mild smoothness condition:

Assume Lipschitz continuity of the conditional expectation for all N ≥ n0: where , CN ∈ O(1) is a positive constant, and we have taken an outer expectation on both sides. Then 𝔼(Y | Ũ) = limN→∞ 𝔼(Y | U, L, B) almost surely.

We delegate proofs to the Supplementary Materials unless proven here in the Methods. Note that , so the Lipschitz assumption intuitively means that accurate estimation of Ũ implies accurate estimation of 𝔼(Y | Ũ). Furthermore, conditioning on the library size N instead of L can introduce spurious dependencies because N depends on all of the genes rather than just the stably expressed ones.

We now eliminate the need to condition on L. Note that L is a sum of independent Poisson distributions given B per Expression (4). This implies Y ╨ L | (U, B) for any N, so that 𝔼(Y |Ũ) = limN→∞ 𝔼(Y | U, L, B) = limN→∞ 𝔼(Y | U, B) almost surely. We have proved:

Consider the same assumption as Lemma 1. Then 𝔼(Y |Ũ) = limN→∞ 𝔼(Y | U, B) almost surely, where we have eliminated the conditioning on L.

We emphasize again that these equalities hold for the conditional expectation but not for the regression parameters; the regression parameters do not converge in general unless we divide by L. We will only need to estimate conditional expectations in order to identify root causal genes.

Identifying Root Causal Genes

We showed how to overcome Poisson measurement error without sequencing depth normalization in the previous section. We leverage this technique to define a measure for identifying the root causal genes of Y.

Definitions

A root cause of Y corresponds to a root vertex that is an ancestor of Y in 𝔾. All root vertices are error terms in an augmented graph. We define the root causal effect of any Ei ∈ E on Y as ϒi ≜ 𝕡(Y | Ei) − 𝕡(Y) (Strobl, 2024; Strobl and Lasko, 2023c).

We can identify root causes using the following result:

If Y or (or both), then Ei is a root cause of Y.

We can also claim the backward direction under d-separation faithfulness. We however avoid making this additional assumption because real biological data may not arise from distributions obeying d-separation faithfulness in practice (Strobl, 2022).

Proposition 1 implies that Ei is a root cause of Y when: The above quantity corresponds to the conditional root causal effect but not the root causal effect ϒi due to the extra conditioning on . The two terms may also differ in direction; if Δi > 0, then this does not imply that ϒi > 0, and similarly for negative values. The two variables thus represent different quantities but – in terms of priority – we would estimate ϒi when we have nonzero Δi. Experimental results indicate that ϒi and Δi take on similar values and agree in direction about 95% of the time in practice (Supplementary Materials).

We now encounter two challenges. First, the quantities ϒi and Δi depend on the unknown error term Ei. We can however substitute Ei with in Δi due to the following result:

We have under Equation (3).

We can thus compute the conditional root causal effect Δi without access to the error terms: We can determine the root causal status of Ei on Y when Δi ≠ 0 per Proposition 1. Nevertheless, the term “root cause” in colloquial language refers to two concepts simultaneously: a root vertex that causes Y and has a large causal effect on Y. We thus say that is a root causal gene of Y if Δi ≫ 0.

The second challenge involves computing the non-parametric probability distributions of Δi which come at a high cost. We thus define the analogous expected version by: where p(Y) denotes the density of Y. Observe that if Δi = 0, then Γi = 0. The converse is not true but likely to hold in real data when a change in the probability distribution also changes its expectation. The set denotes the surrogate parents of corresponding to the variables in X associated with . The last equality holds almost surely as N → ∞ by Theorem 1.

We call Φ ≜ |Γ| the Root Causal Strength (RCS) of on Y. The RCS obtains a unique value Φ = ϕij for each patient j. We say that is a root causal gene of Y for patient if ϕij ≫ 0, since we posit a right skewed distribution of conditional root causal effects for each patient as in Figure 1 (b). We combine the RCS scores across a set of n samples using the Deviation of the RCS (D-RCS) , or the standard deviation of RCS from zero. We may compute D-RCS for each cluster or globally across all patients depending on the context. We thus likewise say that is a root causal gene for a cluster of patients or all patients in a sample if its corresponding D-RCS score for the cluster or the sample is much lager than zero, respectively. Note that we do not specify a particular cutoff value for large (conditional) root causal effects, since the root causal effects likely lie on a continuous graduated scale as opposed to approximately two binary values. Nevertheless, visual inspection of the RCS or D-RCS histograms in disease should approximate a power law, where a large mass is concentrated around zero and a long tail extends to the right similar to folding Figure 1 (b).

Algorithm

We now design an algorithm called Root Causal Strength using Perturbations (RCSP) that recovers the RCS scores using Perturb-seq and bulk RNA-seq data.

Finding Surrogate Ancestors

Computing Φi for each requires access to the surrogate parents of each variable or, equivalently, the causal graph 𝔾. However, inferring 𝔾 using causal discovery algorithms may lead to large statistical errors in the high dimensional setting (Colombo et al., 2014) and require restrictive assumptions such as d-separation faithfulness (Spirtes et al., 2000) or specific functional relations (Peters et al., 2014).

We instead directly utilize the interventional Perturb-seq data to recover a superset of the surrogate parents. We first leverage the global Markov property and equivalently write: where denotes the surrogate ancestors of , or the variables in X associated with the ancestors of .

We discover the surrogate ancestors using unconditional independence tests. For any Xk ∈ X, we test Xk ╨ Pi by unpaired two-sided t-test, where Pi is an indicator function equal to one when we perturb Xi and zero in the control samples of Perturb-seq. Pi is thus a parent of Xi alone but not a child of B, so we do not need to condition on B. We use the two-sided t-test to assess for independence because the t-statistic averages over cells to mimic bulk RNA-seq. If we reject the null and conclude that , then Xk must be a descendant of Pi by the global Markov property, so we include Xk into the set of surrogate descendants . Curating every Xj ∈ X such that into yields the surrogate ancestors of as desired.

Procedure

We now introduce an algorithm called Root Causal Strength using Perturbations (RCSP) that discovers the surrogate ancestors of each variable using Perturb-seq and then computes the RCS of each variable using bulk RNA-seq. We summarize RCSP in Algorithm 1.

RCSP takes Perturb-seq and bulk RNA-seq datasets as input. The algorithm first finds the surrogate descendants of each variable in in Line 2 in order to identify the surrogate ancestors of each variable in Line 5. Access to the surrogate ancestors and the batches B allows RCSP to compute Φi for each Xi ∈ X from the bulk RNA-seq in Line 6. The algorithm thus outputs the RCS scores Φ as desired.

We certify RCSP as follows:

(Fisher consistency) Consider the same assumption as Lemma 1. If unconditional d-separation faithfulness holds, then RCSP recovers Φ almost surely as N → ∞.

Root Causal Strength using Perturbations (RCSP)

We engineered RCSP to only require unconditional d-separation faithfulness because real distributions may not obey full d-separation faithfulness (Strobl, 2022).

Synthetic Data

Simulations

We generated a linear SEM obeying Equation (3) specifically as for every and similarly . We included p + 1 = 2500 variables in . We instantiated the coefficient matrix β by sampling from a Bernoulli(2/(p − 1)) distribution in the upper triangular portion of the matrix. The resultant causal graph thus has an expected neighborhood size of 2. We then randomly permuted the ordering of the variables. We introduced weights into the coefficient matrix by multiplying each entry in β by a weight sampled uniformly from [−1, −0.25] ∪ [0.25, 1]. The error terms each follow a standard Gaussian distribution multiplied by 0.5. We introduced batch effects by drawing each entry of the mapping efficiencies π from the uniform distribution between 10 and 1000 for the bulk RNA-seq, and between 0.1 and 1 for the Perturb-seq. We set and then obtained the corrupted surrogate Xi distributed for each and batch j. We chose Y uniformly at random from the set of vertices with at least one parent and no children. We drew 200 samples for the bulk RNA-seq data to mimic a large but common dataset size. We introduced knockdown perturbations in Perturb-seq by subtracting an offset of two in the softplus function:. We finally drew 200 samples for the control and each perturbation condition to generate the Perturb-seq data. We repeated the above procedure 30 times.

Comparators

We compared RCSP against the following four algorithms:

Additive noise model (ANM) (Peters et al., 2014; Strobl and Lasko, 2023a): performs nonlinear regression of Xi on Pa(Xi) ∪ B and then regresses Y on the residuals E \ Ei to estimate |𝔼(Y | E \ Ei) − 𝔼(Y | X, B) | for each Xi ∈ X. The non-linear regression residuals are equivalent to the error terms assuming an additive noise model.

Linear Non-Gaussian Acyclic Model (LiNGAM) (Peters et al., 2014; Strobl and Lasko, 2022): same as above but performs linear instead of non-linear regression.

CausalCell (Wen et al., 2023): selects the top 50 genes with maximal statistical dependence to Y, and then runs the Peter-Clark (PC) algorithm (Spirtes et al., 2000) using a non-parametric conditional independence test to identify a causal graph among the top 50 genes. The algorithm does not perform root causal inference, so we use ANM as above but condition on the estimated parent sets for the top 50 genes and the ancestors inferred from the Perturb-seq data otherwise.

Univariate regression residuals (Uni Reg): regresses Y on Xi ∪ B and estimates the absolute residuals | Y − 𝔼(Y |Xi, B)| for each Xi ∈ X.

Multivariate regression residuals (Multi Reg): similar to above but instead computes the absolute residuals after regressing Y on (X \ Xi) ∪ B.

The first two methods are state-of-the-art approaches used for root causal discovery. Univariate and multivariate regressions do not distinguish between predictivity and causality, but we included them as sanity checks. We performed all non-linear regressions using multivariate adaptive regression splines to control for the underlying regressor (Friedman, 1991). We also standardized all variables before running the regressions to prevent gaming of the marginal variances in causal discovery (Reisach et al., 2021; Ng et al., 2024). We compared the algorithms on their accuracy in estimating Φ.

Real Data

Quality Control

We downloaded Perturb-seq datasets of retinal pigment epithelial cells from the RPE-1 cell line, and undifferentiated blast cells from the K562 cell line (Replogle et al., 2022). We used the genomewide dataset version for the latter. We downloaded the datasets from the scPerturb database on Zenodo (Green et al., 2022) with the same quality controls as the original paper. Replogle et al. computed adjusted library sizes by equalizing the mean library size of control cells within each batch. Cells with greater than a 2000 or 3000 library size, and less than 25% or 11% mitochondrial RNA were kept, respectively. The parameters were chosen by plotting the adjusted library sizes against the mitochondrial RNA counts and then manually setting thresholds that removed low quality cells likely consisting of ambient mRNA transcripts arising from premature cell lysis or cell death.

We next downloaded bulk RNA-seq datasets derived from patients with age-related macular degeneration (AMD; GSE115828) and multiple sclerosis (MS; GSE137143) (Ratnapriya et al., 2019; Kim et al., 2021). We excluded 10 individuals from the AMD dataset including one with an RNA integrity number of 21.92, five missing an integrity number (all others had an integrity number of less than 10), and four without a Minnesota Grading System score. We kept all samples from the MS dataset derived from CD4+ T cells but filtered out genes with a mean of less than 5 counts as done in the original paper.

We finally kept genes that were present in both the AMD bulk dataset and the RPE-1 Perturbseq dataset, yielding a final count of 513 bulk RNA-seq samples and 247,914 Perturb-seq samples across 2,077 genes. We also kept genes that were present in both the MS bulk dataset and the K562 Perturb-seq dataset, yielding a final count of 137 bulk RNA-seq samples and 1,989,578 Perturb-seq samples across 6,882 genes. We included age and sex as a biological variable as covariates for every patient in both datasets in subsequent analyses.

Evaluation Rationale

We do not have access to the ground truth values of Φ in real data. We instead evaluate the RCSP estimates of Φ using alternative sources of ground truth knowledge. We first assess the accuracy of RCS using the control variable age as follows:

Determine if the RCS values of age identify age as a root cause with large causal effect in diseases that progress over time.

Second, few root causal genes should drive pathogenesis because the effects of a few error terms distribute over many downstream genes. We verify the sparsity of root causal genes as follows:

Determine if the distribution of D-RCS concentrates around zero more than the distribution of the Deviation of Statistical Dependence (D-SD) defined as for each gene where Ωi = |𝔼(Y |Xi, B) − 𝔼(Y |B)| and ωij its value for patient j.

Despite the sparsity of root causal genes, we still expect the root causal genes to correspond to at least some known causes of disease:

Determine if genes with the top D-RCS scores correspond to genes known to cause the disease.

Next, the root causal genes initiate the vast majority of pathogenesis, and we often have knowledge of pathogenic pathways even though we may not know the exact gene expression cascade leading to disease. Intervening on root causal genes should also modulate patient symptoms. We thus further evaluate the accuracy of RCSP using pathway and drug enrichment analyses as follows:

Determine if the D-RCS scores identify known pathogenic pathways of disease in pathway enrichment analysis.

Determine if the D-RCS scores identify drugs that treat the disease.

Finally, complex diseases frequently involve multiple pathogenic pathways that differ between patients. Patients with the same complex disease also respond differently to treatment. We hence evaluate the precision of RCS as follows:

Determine if the patient-specific RCS scores identify subgroups of patients involving different but still known pathogenic pathways.

Determine if the patient-specific RCS scores identify subgroups of patients that respond differently to drug treatment.

In summary, we evaluate RCSP in real data based on its ability to (1) identify age as a known root cause, (2) suggest an omnigenic root causal model, (3) recover known causal genes, (4) find known pathogenic pathways, (5) find drugs that treat the disease, and (6,7) delineate patient subgroups.

Enrichment Analyses

Multivariate adaptive regression splines introduce sparsity, but enrichment analysis performs better with a dense input. We can estimate the conditional expectations of Φ using any general nonlinear regression method, so we instead estimated the expectations using kernel ridge regression equipped with a radial basis function kernel (Shawe-Taylor and Cristianini, 2004). We then computed the D-RCS across all patients for each variable in X. We ran pathway enrichment analysis using the fast gene set enrichment analysis (FGSEA) algorithm (Sergushichev, 2016) with one hundred thousand simple permutations using the D-RCS scores and pathway information from the Reactome database (version 1.86.0) (Fabregat et al., 2017). We likewise performed drug set enrichment analysis with the Drug Signature database (version 1.0) (Yoo et al., 2015). We repeated the above procedures for the D-RCS of all clusters identified by hierarchical clustering via Ward’s method (Ward Jr, 1963).